---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# OSPSuite-R

<!-- badges: start -->

<a href="https://ci.appveyor.com/project/open-systems-pharmacology-ci/OSPSuite-R/branch/develop" class="pkgdown-devel">
<img src="https://ci.appveyor.com/api/projects/status/github/Open-Systems-Pharmacology/OSPSuite-R?branch=develop&amp;svg=true" alt="AppVeyor build status"/></a>
<a href="https://app.codecov.io/gh/Open-Systems-Pharmacology/OSPSuite-R" class="pkgdown-devel">
<img src="https://codecov.io/gh/Open-Systems-Pharmacology/OSPSuite-R/branch/develop/graph/badge.svg" alt="codecov"/></a>

<!-- badges: end -->

# Overview

The **ospsuite-R** package provides the functionality of loading,
manipulating, and simulating the simulations created in the Open Systems
Pharmacology Software tools PK-Sim and MoBi.

-   [Documentation](#documentation)
-   [Installation](#installation)
-   [Usage](#usage)
-   [Known issues](#known-issues)
-   [Code of conduct](#code-of-conduct)
-   [Contribution](#contribution)
-   [Licence](#licence)

# Documentation

If you are reading this on GitHub README, please refer to the [online
documentation](https://www.open-systems-pharmacology.org/OSPSuite-R/)
for more details on the package.

In particular, we would recommend that you read the articles in the
following order:

-   [Get
    Started](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/ospsuite.html)
-   [Loading a simulation and accessing
    entities](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/load-get.html)
-   [Changing parameter and molecule start
    values](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/set-values.html)
-   [Running a
    simulation](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/run-simulation.html)
-   [Efficient
    calculations](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/efficient-calculations.html)
-   [Creating
    individuals](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/create-individual.html)
-   [Population
    simulations](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/create-run-population.html)
-   [PK
    Analysis](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/pk-analysis.html)
-   [Sensitivity
    analysis](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/sensitivity-analysis.html)
-   [Table
    parameters](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/table-parameters.html)
-   [Dimensions and
    Units](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/unit-conversion.html)
-   [Working with data sets and import from
    excel](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/observed-data.html)
-   [Working with `DataCombined`
    class](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/data-combined.html)
-   [Visualizations with
    `DataCombined`](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/data-combined-plotting.html)
    

# Installation

The **ospsuite** package is compatible with R version 3.6.x **AND**
4.x.x and can be used on [Windows](#on-windows) and [Linux](#on-linux)
operating systems.

`ospsuite` is not available on CRAN and also depends on three packages that are 
not available on CRAN. Thus the installation process is a bit unconventional, 
please follow carefully the instructions below.


## On Windows

### Pre-requisites

The package requires additional software installations:


  - Visual C++ Runtime available
    [here](https://aka.ms/vs/16/release/vc_redist.x64.exe)
  - .NET Framework 4.7.2 available
    [here](https://go.microsoft.com/fwlink/?LinkID=863265)

**NB**: These pre-requisites are already installed if the OSP Suite was installed before.


### From GitHub (recommended)

The latest released version of the package can be installed from GitHub
using the `{remotes}` package. The code below will download and install
all the required dependencies.

``` r
install.packages("remotes")
rClrURL <- if (R.Version()$major >= 4) {
  "https://github.com/Open-Systems-Pharmacology/rClr/releases/download/v0.9.2/rClr_0.9.2.zip"
} else {
  "https://github.com/Open-Systems-Pharmacology/rClr/releases/download/v0.9.1-R3/rClr_0.9.1.zip"
}

install.packages(rClrURL,
  repos = NULL,
  type = "binary"
)

remotes::install_github("Open-Systems-Pharmacology/OSPSuite.RUtils@*release")
remotes::install_github("Open-Systems-Pharmacology/TLF-Library@*release")
remotes::install_github("Open-Systems-Pharmacology/OSPSuite-R@*release")
```

### From package archive files

It is also possible to install manually from archive pre-built archive files.

### Install CRAN dependencies

When installing from such files, the CRAN dependencies need to be installed 
manually first.

```{r, echo = FALSE, results="asis"}

# This code chunck gathers all Depends and Imports dependencies of **installed**
# versions of the packages and generate a string that is printed in the README.md
# file.

library(pacman)

non_cran_deps <- c("rClr", "ospsuite.utils", "tlf")
deps <- c()

for (non_cran_dep in non_cran_deps) {
    package_deps <- p_depends(non_cran_dep, local = T, character.only = T)
   deps <- sort(unique(c(deps, package_deps$Depends, package_deps$Imports)))
}

deps <- deps[!deps %in% c("methods", non_cran_deps)]

dep_install <- paste('install.packages("', deps, '")', sep = '', collapse = "\n")



cat("```r \n")
cat("# Install dependencies (e.g. R6) which are on CRAN\n")
cat(dep_install)
cat("\n``` \n")

```

### Install non-CRAN dependencies

Each of the pre-built released packages are available as a a binary `*.zip`.
OSPSuite-R binary archive can be downloaded from
[here](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/releases). The
other non-CRAN dependencies needed for OSPSuite-R also have to be downloaded and
manually installed:

  - `rClr`
      - [For R
        4.x.x](https://github.com/Open-Systems-Pharmacology/rClr/releases/download/v0.9.2/rClr_0.9.2.zip)
      - [For R
        3.6.x](https://github.com/Open-Systems-Pharmacology/rClr/releases/download/v0.9.1-R3/rClr_0.9.1.zip)
  - [`ospsuite.utils`](https://github.com/Open-Systems-Pharmacology/OSPSuite.RUtils/releases/latest)
  - [`tlf`](https://github.com/Open-Systems-Pharmacology/TLF-Library/releases/latest)

If you use [RStudio IDE](https://www.rstudio.com/), you can use the *Install*
option in the *Packages* pane and select the option *Install from -\> Package
Archive File* to install a package from binary `*.zip` files.

**NB**: The CRAN dependencies of rClr, ospuite.utils and tlf were already
installed during the previous step.


```r
# Install `{rClr}` from local file 
# (`pathTo_rCLR.zip` here should be replaced with the actual path to the `.zip` file)
install.packages(pathTo_rCLR.zip, repos = NULL)

# Install `{ospsuite.utils}` from local file 
# (`pathTo_ospsuite.utils.zip` here should be replaced with the actual path to the `.zip` file)
install.packages(pathTo_ospsuite.utils.zip, repos = NULL)

# Install `{tlf}` from local file 
# (`pathTo_tlf.zip` here should be replaced with the actual path to the `.zip` file)
install.packages(pathTo_tlf.zip, repos = NULL)

# Install `{ospsuite}` from local file
# (`pathToOSPSuite.zip` here should be replaced with the actual path to the `.zip` file)
install.packages(pathToOSPSuite.zip, repos = NULL)
```

## On Linux

The **ospsuite** package has been tested under Linux distributions **CentOS 7**
and **Ubuntu 18**. Some functionality, such as creating individuals or
populations, is not available under Linux. Installation under Linux requires
several prerequisites, the detailed instructions can be found in the Wiki:

  - [Setup OSPSuite-R on
    CentOS](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/wiki/Setup-ospsuite-R-on-CentOS7)
  - [Setup OSPSuite-R on
    Ubuntu](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/wiki/Setup-ospsuite-R-on-Ubuntu)

For other Linux distributions Docker containers can be used (Dockerfiles based
on CentOS 7 and Ubuntu 18 are available under
<https://github.com/Open-Systems-Pharmacology/OSPSuite-R/releases>)

## Build from source

You can clone the GIT repository and build the package from source.

### How to update dependencies from nuget?

-   `git submodule update --init --recursive` to install all submodules
-   Make sure you have [ruby](https://www.ruby-lang.org/de/downloads/)
    install and that it is available in your path
-   Run `rake postclean` or simply double click on `postclean.bat`. This
    will update all nuget packages and copy the dependencies in the
    package `inst/lib` folder.


# Known issues

## Loading `ospsuite` might fail if your systems locale is not set to English

``` ibrary(ospsuite)
载入需要的程辑包：rClr
Loading the dynamic library for Microsoft .NET runtime...
Loaded Common Language Runtime version 4.0.30319.42000

Error: package or namespace load failed for ‘ospsuite’:
loadNamespace()里算'ospsuite'时.onLoad失败了，详细内容：
调用: rClr::clrCall(dimensionTask, "AllAvailableUnitNamesFor", enc2utf8(dimension))
错误: Type: System.Collections.Generic.KeyNotFoundException
Message: Dimension 'CV mmHg*s虏/ml' not available in DimensionFactory.
...
```

-- On Windows, set
`Settings > Language > Administrative language settings > Current language for non-Unicode programs`
to `English (United States)` and reboot. -- On Linux, set the
environment variable `LC_ALL` before starting R:

```         
export LC_ALL=en_US.UTF-8
```

## RStudio crashes when trying to load a workspace.

The ospsuite package uses the features implemented in PK-Sim and MoBi by
creating `.NET` objects (e.g. a simulation) and using them from R. These objects
cannot be saved as part of the workspace and reloaded on next start. When trying
to do so, RStudio simply crashes. There is no possibility to overcome this
limitation. To prevent RStudio from crashing, make sure to disable the check-box
"Restore `.RData` into workspace at startup" in the options of RStudio. Keep in
mind that you can also change this setting for specific projects.

# Code of Conduct

Everyone interacting in the Open Systems Pharmacology community
(codebases, issue trackers, chat rooms, mailing lists etc.) is expected
to follow the Open Systems Pharmacology [code of
conduct](https://github.com/Open-Systems-Pharmacology/Suite/blob/master/CODE_OF_CONDUCT.md).

# Contribution

We encourage contribution to the Open Systems Pharmacology community.
Before getting started please read the [contribution
guidelines](https://github.com/Open-Systems-Pharmacology/Suite/blob/master/CONTRIBUTING.md).
If you are contributing to the codebase, please be familiar with the [coding
standards](https://github.com/Open-Systems-Pharmacology/developer-docs/blob/main/ospsuite-r-specifics/CODING_STANDARDS_R.md).

# License

OSPSuite-R is released under the [GPLv2 License](LICENSE).

All trademarks within this document belong to their legitimate owners.
