---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# OSPSuite-R

<!-- badges: start -->
[![Main-Workflow](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/actions/workflows/main-workflow.yaml/badge.svg)](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/actions/workflows/main-workflow.yaml)
[![Codecov test coverage](https://codecov.io/gh/Open-Systems-Pharmacology/OSPSuite-R/branch/main/graph/badge.svg)](https://app.codecov.io/gh/Open-Systems-Pharmacology/OSPSuite-R?branch=main)
<!-- badges: end -->

# Overview

The **ospsuite-R** package provides the functionality of loading,
manipulating, and simulating the simulations created in the Open Systems
Pharmacology Software tools PK-Sim and MoBi.

-   [Documentation](#documentation)
-   [Installation](#installation)
-   [Known issues](#known-issues)
-   [Code of conduct](#code-of-conduct)
-   [Contribution](#contribution)
-   [Licence](#licence)

# Documentation

If you are reading this on GitHub README, please refer to the [online
documentation](https://www.open-systems-pharmacology.org/OSPSuite-R/)
for more details on the package.

In particular, we would recommend that you read the articles in the
following order:

-   [Get
    Started](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/ospsuite.html)
-   [Loading a simulation and accessing
    entities](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/load-get.html)
-   [Changing parameter and molecule start
    values](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/set-values.html)
-   [Running a
    simulation](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/run-simulation.html)
-   [Efficient
    calculations](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/efficient-calculations.html)
-   [Creating
    individuals](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/create-individual.html)
-   [Population
    simulations](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/create-run-population.html)
-   [PK
    Analysis](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/pk-analysis.html)
-   [Sensitivity
    analysis](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/sensitivity-analysis.html)
-   [Table
    parameters](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/table-parameters.html)
-   [Dimensions and
    Units](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/unit-conversion.html)
-   [Working with data sets and import from
    excel](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/observed-data.html)
-   [Working with `DataCombined`
    class](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/data-combined.html)
-   [Visualizations with
    `DataCombined`](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/data-combined-plotting.html)
    

# Installation

The **ospsuite** package is compatible with R version 4.x.x and can be used on 
[Windows](#on-windows) and [Linux (Ubuntu)](#on-linux) operating systems.

`ospsuite` is not available on CRAN and also depends on packages from the OSP ecosystem
that are not available on CRAN. Please follow the instructions below to install
the packages and all required dependencies.


## On Windows

### Pre-requisites

The package requires additional software installations:


- Latest Microsoft Visual C++ Redistributable for Visual Studio 2015, 2017, 2019 and 2022 available [here](https://support.microsoft.com/en-us/help/2977003/the-latest-supported-visual-c-downloads) 
- .NET 8 runtime available [here](https://dotnet.microsoft.com/download/dotnet/8.0/runtime).

**NB**: These pre-requisites are already installed if the OSP Suite was installed before.

### From GitHub (recommended)

The latest released version of the package can be installed from GitHub
using the `{pak}` package. The code below will download and install
all the required dependencies.

```{r, eval=FALSE}
install.packages("pak")
pak::pak("Open-Systems-Pharmacology/OSPSuite-R@*release")
```

Get the latest development version with:

```{r, eval=FALSE}
install.packages("pak")
pak::pak("Open-Systems-Pharmacology/OSPSuite-R")
```

### From package archive files

It is also possible to install manually from archive pre-built archive files provided
with the [release](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/releases).

#### Install CRAN dependencies

When installing from such files, the CRAN dependencies need to be installed 
manually first.

```{r, echo = FALSE, results="asis"}
# This code chunck gathers all Depends and Imports dependencies of **installed**
# versions of the packages and generate a string that is printed in the README.md
# file.

library(pacman)

non_cran_deps <- c("rSharp", "ospsuite.utils", "tlf", "ospsuite")
deps <- c()

for (non_cran_dep in non_cran_deps) {
  package_deps <- p_depends(non_cran_dep, local = T, character.only = T)
  deps <- sort(unique(c(deps, package_deps$Depends, package_deps$Imports)))
}

deps <- deps[!deps %in% c("methods", non_cran_deps)]

dep_install <- paste('install.packages("', deps, '")', sep = "", collapse = "\n")



cat("```r \n")
cat("# Install dependencies (e.g. R6) which are on CRAN\n")
cat(dep_install)
cat("\n``` \n")
```

#### Install non-CRAN dependencies

Each of the pre-built released packages are available as a a binary `*.zip`.
OSPSuite-R binary archive can be downloaded from
[here](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/releases). The
other non-CRAN dependencies needed for OSPSuite-R also have to be downloaded and
manually installed:

  - [`rSharp`](https://github.com/Open-Systems-Pharmacology/rSharp/releases/latest)
  - [`ospsuite.utils`](https://github.com/Open-Systems-Pharmacology/OSPSuite.RUtils/releases/latest)
  - [`tlf`](https://github.com/Open-Systems-Pharmacology/TLF-Library/releases/latest)

If you use [RStudio IDE](https://www.rstudio.com/), you can use the *Install*
option in the *Packages* pane and select the option *Install from -\> Package
Archive File* to install a package from binary `*.zip` files.

**NB**: The CRAN dependencies of rSharp, ospuite.utils and tlf were already
installed during the previous step.


```r
# Install `{rSharp}` from local file 
# (`pathTo_rSharp.zip` here should be replaced with the actual path to the `.zip` file)
install.packages(pathTo_rSharp.zip, repos = NULL)

# Install `{ospsuite.utils}` from local file 
# (`pathTo_ospsuite.utils.zip` here should be replaced with the actual path to the `.zip` file)
install.packages(pathTo_ospsuite.utils.zip, repos = NULL)

# Install `{tlf}` from local file 
# (`pathTo_tlf.zip` here should be replaced with the actual path to the `.zip` file)
install.packages(pathTo_tlf.zip, repos = NULL)

# Install `{ospsuite}` from local file
# (`pathToOSPSuite.zip` here should be replaced with the actual path to the `.zip` file)
install.packages(pathToOSPSuite.zip, repos = NULL)
```

## On Linux

The **ospsuite** package has been tested under Linux distribution
**Ubuntu 22.04**. Installation
under Linux requires several prerequisites, the detailed instructions
can be found in the Wiki:

- [Setup OSPSuite-R on
  Ubuntu](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/wiki/Setup-ospsuite-R-on-Ubuntu)

For other Linux distributions Docker containers can be used (Docker container 
based on Ubuntu 22 is available under
<https://github.com/Open-Systems-Pharmacology/OSPSuite-R/releases>)

## Build from source

You can clone the GIT repository and build the package from source.

### How to update the libraries?

The `{ospsuite}` package requires some shared libraries to get access to PK-Sim 
functionality. To get the latest libraries(**.dll** on Windows or **.so** on Linux), run
the script file `update_core_files.R' provided with this package.


# Known issues

## Loading `ospsuite` might fail if your systems locale is not set to English

-- On Windows, set
`Settings > Language > Administrative language settings > Current language for non-Unicode programs`
to `English (United States)` and reboot. -- On Linux, set the
environment variable `LC_ALL` before starting R:

```         
export LC_ALL=en_US.UTF-8
```

## Saving and loading the workspace in RStudio does not restore objects

The ospsuite package uses the features implemented in PK-Sim and MoBi by
creating `.NET` objects (e.g. a simulation) and using them from R. These
objects cannot be saved as part of the workspace and reloaded on next
start. Upon restoring the workspace, the objects will be `NULL` and cannot be
re-used.

# Code of Conduct

Everyone interacting in the Open Systems Pharmacology community
(codebases, issue trackers, chat rooms, mailing lists etc.) is expected
to follow the Open Systems Pharmacology [code of
conduct](https://github.com/Open-Systems-Pharmacology/Suite/blob/master/CODE_OF_CONDUCT.md).

# Contribution

We encourage contribution to the Open Systems Pharmacology community.
Before getting started please read the [contribution
guidelines](https://github.com/Open-Systems-Pharmacology/Suite/blob/master/CONTRIBUTING.md).
If you are contributing to the codebase, please be familiar with the [coding
standards](https://github.com/Open-Systems-Pharmacology/developer-docs/blob/main/ospsuite-r-specifics/CODING_STANDARDS_R.md).

# License

OSPSuite-R is released under the [GPLv2 License](LICENSE).

All trademarks within this document belong to their legitimate owners.
