---
output: github_document
---

``` {r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# OSPSuite R package

<!-- badges: start -->

[![Build Badge](https://img.shields.io/github/actions/workflow/status/Open-Systems-Pharmacology/OSPSuite-R/main-workflow.yaml?branch=main&label=Build)](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/actions/workflows/main-workflow.yaml)
[![Codecov test coverage Badge](https://codecov.io/gh/Open-Systems-Pharmacology/OSPSuite-R/branch/main/graph/badge.svg)](https://app.codecov.io/gh/Open-Systems-Pharmacology/OSPSuite-R?branch=main)

<!-- badges: end -->

# Overview

The `{ospsuite}`R package provides the functionality of loading, manipulating,
and running the simulations created in the Open Systems Pharmacology Software
tools PK-Sim and MoBi.

  - [Documentation](#documentation)
  - [Installation](#installation)
  - [Known issues](#known-issues)
  - [Development](#development)
  - [Code of conduct](#code-of-conduct)
  - [Contribution](#contribution)
  - [Licence](#licence)

# Documentation

For complete documentation, see the [online
documentation](https://www.open-systems-pharmacology.org/OSPSuite-R/).

## Essential Guides

**Essential guides:**

  - [Get Started](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/ospsuite.html) - Complete beginner tutorial with working examples
  - [Loading a simulation and accessing entities](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/load-get.html)
  - [Changing parameter and molecule start values](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/set-values.html)
  - [Running a simulation](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/run-simulation.html)

**For advanced workflows:**

  - [Efficient calculations](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/efficient-calculations.html)
  - [Creating individuals](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/create-individual.html)
  - [Population simulations](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/create-run-population.html)
  - [PK Analysis](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/pk-analysis.html)
  - [Sensitivity analysis](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/sensitivity-analysis.html)

**Data handling and visualization:**

  - [Working with data sets and import from excel](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/observed-data.html)
  - [Working with `DataCombined` class](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/data-combined.html)
  - [Visualizations with `DataCombined`](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/data-combined-plotting.html)

**Reference:**

  - [Dimensions and Units](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/unit-conversion.html)
  - [Table parameters](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/table-parameters.html)
  - [PK-Sim Installation](https://www.open-systems-pharmacology.org/OSPSuite-R/articles/PKSim-installation.html)


# Installation

The **ospsuite** package requires R version 4.x.x and supports
[Windows](#on-windows) and [Linux (Ubuntu)](#on-linux) operating systems.

`ospsuite` is not available on CRAN and depends on packages from the OSP
ecosystem that are not available on CRAN. Follow the instructions below
to install the package and all required dependencies.


## Pre-requisites

The `{ospsuite}` package requires `{rSharp}` and its external dependencies (Visual
C++ Redistributable and .NET 8). Install these dependencies using the following instructions:

  - [For
    Windows](https://github.com/Open-Systems-Pharmacology/rSharp?tab=readme-ov-file#prerequisites)
  - [For
    Linux](https://github.com/Open-Systems-Pharmacology/rSharp?tab=readme-ov-file#ubuntu)

## From GitHub (recommended)

Install the latest released version from GitHub using
the `{remotes}` package. This installation method downloads and installs all
required dependencies.

```{r, eval=FALSE}
install.packages("remotes")
remotes::install_github("Open-Systems-Pharmacology/OSPSuite-R@*release")
```

To install the latest development version:

```{r, eval=FALSE}
install.packages("remotes")
remotes::install_github("Open-Systems-Pharmacology/OSPSuite-R")
```

## From package archive files (Deprecated)

Install manually from pre-built archive files available in the [releases](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/releases).

#### Install CRAN dependencies

Installing from archive files requires manual installation of CRAN dependencies first.

```{r, echo = FALSE, results="asis"}
# This code chunck gathers all Depends and Imports dependencies of **installed**
# versions of the packages and generate a string that is printed in the README.md
# file.

library(pacman)

non_cran_deps <- c("rSharp", "ospsuite.utils", "tlf", "ospsuite")
deps <- c()

for (non_cran_dep in non_cran_deps) {
  package_deps <- p_depends(non_cran_dep, local = T, character.only = T)
  deps <- sort(unique(c(deps, package_deps$Depends, package_deps$Imports)))
}

deps <- deps[!deps %in% c("methods", non_cran_deps)]

dep_install <- paste(
  'install.packages("',
  deps,
  '")',
  sep = "",
  collapse = "\n"
)


cat("```r \n")
cat("# Install dependencies (e.g. R6) which are on CRAN\n")
cat(dep_install)
cat("\n``` \n")
```

#### Install non-CRAN dependencies

Pre-built packages are available as binary files (*.zip for Windows, *.tar.gz for Linux)
Download the OSPSuite-R binary archive from the
[releases page](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/releases). Download and install these additional non-CRAN dependencies:

  - [`{rSharp}`](https://github.com/Open-Systems-Pharmacology/rSharp/releases/latest)
  - [`{ospuite.utils}`](https://github.com/Open-Systems-Pharmacology/OSPSuite.RUtils/releases/latest)
  - [`{tlf}`](https://github.com/Open-Systems-Pharmacology/TLF-Library/releases/latest)

In [RStudio IDE](https://www.rstudio.com/), use the *Install*
option in the *Packages* pane and select *Install from -> Package
Archive File* to install packages from binary files.

**Note**: CRAN dependencies for `{rSharp}`, `{ospuite.utils}`, and `{tlf}` were installed in the previous step.


``` r
# Install rSharp from local file
# Replace pathTo_rSharp.zip with the actual path to the .zip file (or to the .tar.gz file on Linux)
install.packages(pathTo_rSharp.zip, repos = NULL)

# Install ospsuite.utils from local file
# Replace pathTo_ospsuite.utils.zip with the actual path to the .zip file (or to the .tar.gz file on Linux)
install.packages(pathTo_ospsuite.utils.zip, repos = NULL)

# Install tlf from local file
# Replace pathTo_tlf.zip with the actual path to the .zip file (or to the .tar.gz file on Linux)
install.packages(pathTo_tlf.zip, repos = NULL)

# Install ospsuite from local file
# Replace pathToOSPSuite.zip with the actual path to the .zip file (or to the .tar.gz file on Linux)
install.packages(pathTo_OSPSuite.zip, repos = NULL)

```

# Known issues

## Loading `ospsuite` might fail if the system locale is not set to English

  - On Windows, set `Settings > Language > Administrative language settings > Current language for non-Unicode programs`
  to `English (United States)` and reboot.

  - On Linux, set the environment variable `LC_ALL` to `~/.bashrc` before starting R:

    `export LC_ALL=en_US.UTF-8`

  -  add the path to the lib subfolder of installed ospsuite package to LD_LIBRARY_PATH
    `export LD_LIBRARY_PATH=/usr/local/lib/R/site-library/ospsuite/lib/:$LD_LIBRARY_PATH`

## Saving and loading the workspace in RStudio does not restore objects

The ospsuite package uses the features implemented in PK-Sim and MoBi by
creating `.NET` objects (e.g. a simulation) and using them from R. These objects
cannot be saved as part of the workspace and reloaded on next start. Upon
restoring the workspace, the objects will be `NULL` and cannot be re-used.

# Development

## Embedded binaries

The `{ospsuite}` package requires shared libraries (binary files) to access PK-Sim and MoBi
functionalities.

**Note**: Because `{ospsuite}` contains binary files, it is classified as a binary
package and cannot be submitted to CRAN.

### Updating embedded binaries

The embedded binaries are built from NuGet packages specified in the DependencyManager project. To update the binaries to a new version:

1.  Open `shared/DependencyManager/src/DependencyManager.csproj`.
2.  Update the `Version` attribute in the `PackageReference` elements to the desired version numbers.
3.  Commit and push the changes to a branch.
4.  The GitHub Actions workflow defined in `.github/workflows/build-libraries.yaml` automatically:
    - Builds the DependencyManager project, which downloads the specified NuGet packages
    - Extracts the required binary files and dependencies
    - Copies them to the `inst/lib/` directory
    - Commits the updated binaries to the branch

No manual build steps are required. The workflow handles the entire process automatically when changes to the `.csproj` file are pushed.

**Note:** If the `DependencyManager.csproj` file is not modified, the build-libraries workflow step is automatically skipped.

## Versioning

The package follows the versioning process described in the [OSP R collaboration
guide](https://dev.open-systems-pharmacology.org/r-development-resources/collaboration_guide#releasing-versions).

For development versions, [this GitHub
action](https://github.com/Open-Systems-Pharmacology/OSPSuite-R/blob/main/.github/workflows/main-workflow.yaml#L11-L17)
automatically increments the `.9000` version suffix when pull requests are merged.


# Code of Conduct

Everyone interacting in the Open Systems Pharmacology community (codebases,
issue trackers, chat rooms, mailing lists etc.) is expected to follow the Open
Systems Pharmacology [code of
conduct](https://github.com/Open-Systems-Pharmacology/Suite/blob/master/CODE_OF_CONDUCT.md).

# Contribution

Contributions to the Open Systems Pharmacology community are welcome. Before
contributing, read the [contribution
guidelines](https://github.com/Open-Systems-Pharmacology/Suite/blob/master/CONTRIBUTING.md).
Code contributors should follow the [R coding
standards](https://dev.open-systems-pharmacology.org/r-development-resources/coding_standards_r)
and [collaboration
guide](https://dev.open-systems-pharmacology.org/r-development-resources/collaboration_guide).

# License

OSPSuite-R is released under the [GPLv2 License](LICENSE).

All trademarks within this document belong to their legitimate owners.
