---
title: "Creating individuals"
output: rmarkdown::html_vignette
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Creating individuals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The ospsuite-R package provides an interface to the PK-Sim physiology database to create parameter sets describing a certain **individual**. By applying these parameter values to a simulation, it is possible to simulate different individuals based on one exported *.pkml simulation. 
This functionality is only available when PK-Sim is installed on the system.

The easier way to get started is to install the OSPSuite Setup. This will link PK-Sim to your ospsuite-R installation automatically. 
If the portable version of PK-Sim is used instead,  one must specify the path to the PK-Sim folder by calling the function `initPKSim`:

## Creating individuals
The physiology of an individual is defined by the values of the simulation parameters. To simulate a specific individual, one has to generate a set of parameter values and apply these values to the model. **NOTE**: Currently, only individuals of the *same* species as in the original *.pkml simulation can be applied. I.e., if the simulation exported from PK-Sim represents a human individual, it is possible to simulate another human individual of different race, gender, etc., but it is not possible to simulate a rat. Though it is technically possible, and the simulation will produce some results, these will not valid.

The first step is creating an object describing *individual characteristics*. To see the list of available values for the arguments `species`, `population` (only for human), and `gender` (only for human), use the enums `Species`, `HumanPopulation`, and `Gender`, respectively. This object is then passed to the function `createIndividual` to generate a set of parameter values. The algorithm behind is the same used in PK-Sim when creating an Individual-Building Block.

```{r createIndividualCharacteristics}
library(ospsuite)

#If no unit is specified, the default units are used. For "weight" it is "kg", for "age" it is "year(s)".
individualCharacterstics <- createIndividualCharacteristics(species = Species$Human,
                                                            population = HumanPopulation$Japanese_Population,
                                                            gender = Gender$Female,
                                                            weight = 75,
                                                            height = 1.75,
                                                            heightUnit = "m",
                                                            age = 43)
print(individualCharacterstics)

individual <- createIndividual(individualCharacteristics = individualCharacterstics)
```

The output contains two lists of parameters - `distributedParameters` and `derivedParameters`. The first list contains parameters that differ between the individuals of the selected species, while the second list mostly consists of values of parameters defined by formulas in the simulation. When applying the generated individual parameter set to a simulation in R, only parameters from the `distributedParameters` should be overwritten, otherwise formula dependencies may be destroyed. Generated parameter values can be conveniently applied using the `setParameterValuesByPath` method:

```{r setIndividualParameters}
library(ospsuite)

# Load simulation
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
print(simFilePath)
sim <- loadSimulation(simFilePath)

# Apply individual parameters
setParameterValuesByPath(parameterPaths = individual$distributedParameters$paths,
                         values = individual$distributedParameters$values,
                         simulation = sim)
```

## Adding enzyme ontogenies
The PK-Sim database includes ontogeny information for some proteins (see (PK-Sim Ontogeny Database)[https://github.com/Open-Systems-Pharmacology/OSPSuite.Documentation/blob/master/PK-Sim%20Ontogeny%20Database%20Version%207.3.pdf]). For a protein molecule present in the simulation, it is possible to add the ontogeny information on one of the pre-defined proteins. For example, it is possible to set the ontogeny of the protein `MyProtein` to the value of the ontogeny of a CYP3A4 enzyme for the specified individual. The list of supported ontogenies is stored in the `StandardOntogeny`-list.

```{r addOntogeny}
library(ospsuite)

# All supported ontogenies
print(StandardOntogeny)

#Create the ontogeny for the protein "MyProtein" based on ontology of CYP3A4
myProteinOntogeny <- MoleculeOntogeny$new(molecule = "MyProtein", ontogeny = StandardOntogeny$CYP3A4)
print(myProteinOntogeny)

# Add this ontogeny to the individual characteristics used to create the individual parameters set
individualCharacterstics <- createIndividualCharacteristics(species = Species$Human,
                                                            population = HumanPopulation$Japanese_Population,
                                                            gender = Gender$Female,
                                                            weight = 75,
                                                            height = 1.75,
                                                            heightUnit = "m",
                                                            age = 43,
                                                            moleculeOntogenies = myProteinOntogeny)
print(individualCharacterstics)

individual <- createIndividual(individualCharacteristics = individualCharacterstics)
```
