---
title: "Visualizations with `DataCombined`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizations with `DataCombined`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  dpi = 300,
  comment = "#>",
  out.width = "100%",
  fig.height = 6,
  fig.width = 8
)
```

## Introduction

You have already seen how the `DataCombined` class can be utilized to store observed and/or simulated data (if not, read [Working with `DataCombined` class](data-combined.html)).

Let's first create a `DataCombined` object, which we will use to demonstrate different visualizations available.

```{r}
library(ospsuite)

# simulated data
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)
simResults <- runSimulation(sim)
outputPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"

# observed data
obsData <- lapply(
  c("ObsDataAciclovir_1.pkml", "ObsDataAciclovir_2.pkml", "ObsDataAciclovir_3.pkml"),
  function(x) loadDataSetFromPKML(system.file("extdata", x, package = "ospsuite"))
)
names(obsData) <- lapply(obsData, function(x) x$name)

myDataCombined <- DataCombined$new()

myDataCombined$addSimulationResults(
  simulationResults = simResults,
  quantitiesOrPaths = outputPath,
  groups = "Aciclovir PVB"
)

myDataCombined$addDataSets(
  obsData$`Vergin 1995.Iv`,
  groups = "Aciclovir PVB"
)
```

## Profile plots

```{r}
plotIndividualTimeProfile(myDataCombined)
```

## Observed versus simulated scatter plot 

```{r}
plotObservedVsSimulated(myDataCombined)
```

## Residuals versus simulated scatter plot

```{r}
plotResidualsVsSimulated(myDataCombined)
```

## Residuals versus time scatter plot

```{r}
plotResidualsVsTime(myDataCombined)
```

## Customizing plot aesthetics

The aesthetics for plots can be customized using the `DefaultPlotConfiguration` class, which provides various class members that can be used to modify the *look* of the plot.

```{r}
myPlotConfiguration <- DefaultPlotConfiguration$new()

myPlotConfiguration$xUnit <- ospUnits$Time$s
myPlotConfiguration$yUnit <- ospUnits$`Concentration [mass]`$`µg/l`

myPlotConfiguration$title <- "Example: Customizing a Plot"
myPlotConfiguration$subtitle <- "Using `DefaultPlotConfiguration` class"
myPlotConfiguration$caption <- "Source: `Aciclovir` data"

myPlotConfiguration$legendPosition <- tlf::LegendPositions$outsideRight
myPlotConfiguration$yAxisScale <- tlf::Scaling$log
```

This configuration class can be passed to all plotting functions:

```{r}
plotIndividualTimeProfile(myDataCombined, myPlotConfiguration)
```
