---
title: "Visualizations with `DataCombined`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualizations with `DataCombined`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R")
knitr::opts_chunk$set(
  eval = .ospRuntimeAvailable,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  out.width = "100%",
  fig.height = 6,
  fig.width = 8,
  fig.showtext = TRUE
)
```

## Introduction

You have already seen how the `DataCombined` class can be used to store observed
and/or simulated data (if not, read [Working with `DataCombined`](data-combined.html)).

This vignette covers the plotting functions that work directly with `DataCombined`
objects. All functions return standard `ggplot2` objects that can be further
customised with the usual `ggplot2` layers. For a full reference of all available
plot types and parameters see [Plotting with ospsuite.plots](plotting-with-ospsuite-plots.html).

```{r setup}
library(ospsuite)
```

Let's create a `DataCombined` object to use throughout this vignette.

```{r data-setup}
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)
simResults <- runSimulations(sim)[[1]]
outputPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"

obsData <- lapply(
  c("ObsDataAciclovir_1.pkml", "ObsDataAciclovir_2.pkml", "ObsDataAciclovir_3.pkml"),
  function(x) loadDataSetFromPKML(system.file("extdata", x, package = "ospsuite"))
)
names(obsData) <- lapply(obsData, function(x) x$name)

myDataCombined <- DataCombined$new()

myDataCombined$addSimulationResults(
  simulationResults = simResults,
  quantitiesOrPaths = outputPath,
  groups = "Aciclovir PVB"
)

myDataCombined$addDataSets(
  obsData$`Vergin 1995.Iv`,
  groups = "Aciclovir PVB"
)
```

## Time profile plots

Time profile plots visualise measured or simulated values against time and help
assess whether observed data (symbols and error bars) match simulated data (lines).

```{r time-profile}
plotTimeProfile(myDataCombined)
```

## Predicted versus observed scatter plot

Predicted versus observed plots assess how far simulated results are from
observed values.

```{r pred-vs-obs}
plotPredictedVsObserved(myDataCombined)
```

The identity line represents perfect agreement. By default a two-fold range is
shown as dashed lines. Pass additional fold distances via `comparisonLineVector`:

```{r pred-vs-obs-fold}
plotPredictedVsObserved(
  myDataCombined,
  comparisonLineVector = ospsuite.plots::getFoldDistanceList(folds = c(1.3, 2))
)
```

## Residuals versus covariate scatter plots

Residual plots reveal systematic bias in simulated values across time or
concentration regions.

### Residuals versus simulated values

```{r res-vs-pred}
plotResidualsVsCovariate(myDataCombined, xAxis = "predicted")
```

### Residuals versus time

```{r res-vs-time}
plotResidualsVsCovariate(myDataCombined, xAxis = "time")
```

The `residualScale` argument controls how residuals are computed and displayed
(`"log"` by default, or `"linear"` / `"ratio"`):

```{r res-vs-time-linear}
plotResidualsVsCovariate(myDataCombined, xAxis = "time", residualScale = "linear")
```

## Customizing plots

All functions return `ggplot2` objects, so any standard `ggplot2` layer can be
added directly.

### Units and axis scaling

Pass `xUnit` / `yUnit` to set the display units, and `yScale` to change the
axis scale:

```{r custom-units}
plotTimeProfile(
  myDataCombined,
  xUnit = ospUnits$Time$s,
  yUnit = ospUnits$`Concentration [mass]`$`µg/l`,
  yScale = "log"
)
```

### Title, subtitle, and caption

```{r custom-labels}
plotTimeProfile(myDataCombined) +
  ggplot2::labs(
    title = "Aciclovir — Individual Time Profile",
    subtitle = "Simulated vs. Observed (Vergin 1995)",
    caption = "Source: Aciclovir data set"
  )
```

### Legend position

```{r custom-legend}
plotPredictedVsObserved(myDataCombined) +
  ggplot2::theme(legend.position = "bottom")
```

To place the legend inside the panel, anchored to the top-right corner:

```{r custom-legend-inside, eval = FALSE}
plotTimeProfile(myDataCombined) +
  ggplot2::theme(
    legend.position = "inside",
    legend.position.inside = c(0.95, 0.95),
    legend.justification.inside = c("right", "top")
  )
```

### Further theming

Any `ggplot2::theme()` element can be overridden:

```{r custom-theme}
plotTimeProfile(myDataCombined) +
  ggplot2::theme(
    axis.title = ggplot2::element_text(size = 13, face = "bold"),
    axis.text  = ggplot2::element_text(size = 11)
  )
```

## Creating multi-panel figures

Because each function returns a `ggplot2` object, panels can be assembled with
the `patchwork` package:

```{r multi-panel, fig.width = 10, fig.asp = 1, eval = .ospRuntimeAvailable && requireNamespace("patchwork", quietly = TRUE)}
library(patchwork)

p1 <- plotTimeProfile(myDataCombined) +
  ggplot2::labs(tag = "a")
p2 <- plotPredictedVsObserved(myDataCombined) +
  ggplot2::labs(tag = "b")
p3 <- plotResidualsVsCovariate(myDataCombined, xAxis = "predicted") +
  ggplot2::labs(tag = "c")
p4 <- plotResidualsVsCovariate(myDataCombined, xAxis = "time") +
  ggplot2::labs(tag = "d")

(p1 | p2) / (p3 | p4)
```

Control the layout with `patchwork::plot_layout()`:

```{r multi-panel-1col, fig.width = 8, fig.asp = 2.2, eval = .ospRuntimeAvailable && requireNamespace("patchwork", quietly = TRUE)}
p1 / p2 / p3 / p4
```

## Saving plots

Use `ospsuite.plots::exportPlot()` to save a plot to disk:

```{r save-exportplot, eval = FALSE}
plotObject <- plotTimeProfile(myDataCombined)

ospsuite.plots::exportPlot(
  plotObject = plotObject,
  filePath   = "timeprofile.png",
  width      = 8,
  height     = NULL, # auto-computed from content
  dpi        = 300
)
```

Or use `ggplot2::ggsave()` directly:

```{r save-ggsave, eval = FALSE}
ggplot2::ggsave("timeprofile.png", plotObject, width = 8, height = 6, dpi = 300)
```
