---
title: "Working with `DataCombined` class"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with `DataCombined` class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

In Modeling and Simulation workflow, we often need to work multiple observed and/or simulated datasets. Further, some of these datasets can be clubbed together in a group for analysis and visualization.

The `DataCombined` class in `{ospsuite}` provides a container to store these datasets, to group them, and to transform them further.

Before we see how, let's first create an observed and a simulated dataset.

```{r}
library(ospsuite)

# simulated data
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)
simResults <- runSimulation(sim)
outputPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"

# observed data
obsData <- lapply(
  c("ObsDataAciclovir_1.pkml", "ObsDataAciclovir_2.pkml", "ObsDataAciclovir_3.pkml"),
  function(x) loadDataSetFromPKML(system.file("extdata", x, package = "ospsuite"))
)
names(obsData) <- lapply(obsData, function(x) x$name)
```

## Creating `DataCombined` object

First, we create a new instance of `DataCombined` class.

```{r}
myDataCombined <- DataCombined$new()
```

Then we add simulated results to this object using `$addSimulationResults()`:

```{r}
myDataCombined$addSimulationResults(
  simulationResults = simResults,
  quantitiesOrPaths = outputPath,
  names = "Aciclovir Plasma",
  groups = "Aciclovir PVB"
)
```

Next we add observed data to this object using `$addDataSets()`:

```{r}
myDataCombined$addDataSets(
  obsData$`Vergin 1995.Iv`, 
  groups = "Aciclovir PVB"
)
```

There are a few things to keep in mind here:

- It doesn't matter the order in which observed or simulated datasets are added.
- You can add multiple `DataSet` objects in `$addDataSets()` method call.
- You can add only a single instance of `SimulationResults` in `$addSimulationResults()` method call.
- If you add a dataset and then add it one more time, the latter will replace the former. `DataCombined` uses name of the dataset as an identifier to check for uniqueness. 

## Renaming

Since the dataset names can be quite long, sometimes you may wish to rename them. Both `$addDataSets()` and `$addSimulationResults()` allow this via `names` argument. In case of `SimulationResults`, the new names will be for quantities/paths.

Given that datsets names act as unique identifier for these entities, duplicate names are not allowed.

## Grouping

Since the `DataCombined` object can store many datasets, some of them may naturally form a grouping and you would wish to track this in the object. Both `$addDataSets()` and `$addSimulationResults()` allow group specification via `groups` argument.

If you do not specify `groups` when you add datasets or set it to `NULL`, and wish to update groupings later, you can use the `$setGroups()` method.

At any point, you can check the current names and groupings with the following active field:

```{r}
myDataCombined$groupMap
```

## Extracting a combined data frame

The data frame (also sometimes called as a table) data structure is central to R-based workflows, and, thus, we may wish to extract a data frame for datasets present in the object.

Internally, `DataCombined` extracts data frames for observed and simulated datasets and combines them.

```{r}
myDataCombined$toDataFrame()
```

This function returns a [tibble data frame](https://r4ds.had.co.nz/tibbles.html#tibbles-vs.-data.frame). If you wish to modify how it is printed, you can have a look at the available options [here](https://pillar.r-lib.org/reference/pillar_options.html). In fact, let's change a few options and print the data frame again.

```{r, eval=FALSE}
options(
  pillar.width = Inf, # show all columns
  pillar.min_chars = Inf # to turn off truncation of column titles
)

myDataCombined$toDataFrame()
```


```{r, echo=FALSE}
# change these settings only temporarily
withr::with_options(
  list(
    pillar.width = Inf, # show all columns
    pillar.min_chars = Inf # to turn off truncation of column titles
  ),
  code = {
    print(myDataCombined$toDataFrame())
  }
)
```

As can be seen, this data frame attemtps to provide a one-stop access to all important details about the datasets included in the object, including important meta-data.

`{ospsuite}` also provides a few helper functions to modify the data frame further.

When multiple (observed and/or simulated) datasets are present in a data frame, they are likely to have different units. `convertUnits()` function helps to convert them to a common unit^[Note that if you are using `DataCombined` object for plotting functions, you don't need to this conversion; the plotting functions will take care of this internally.].

```{r}
convertUnits(
  myDataCombined,
  xUnit = ospUnits$Time$s,
  yUnit = ospUnits$`Concentration [mass]`$`µg/l`
)
```

Furthermore, when both observed and simulated datasets are present, we can also compute residuals using `calculateResiduals()`:

```{r}
calculateResiduals(myDataCombined, scaling = tlf::Scaling$lin)
```

## Visualizations with `DataCombined`

Plotting the stored data is an important part of model diagnostics and quality control.

See [Visualizations with `DataCombined`](data-combined-plotting.html) describing functions that can visualize data stored in `DataCombined` object.
