---
title: "Efficient calculation"
output: rmarkdown::html_vignette
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Efficient calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

There are a few ways to run simulations using this package. 
The easiest way is, as expected, not necessarily the most efficient one. So depending on the use case, a different approach than what is outlined in [Run Simulations](run-simulation.html) might be more suited.

This vignettes aims at describing some scenarios and the suggested methods to efficiently run simulations.

## Life cycle of a simulation run

In general, a user loads a simulation, update some parameter values, calculates the simulation and retrieves some results. 
Here are some steps inherent to each simulation:

1. Loading a simulation
2. Optional: Setting parameter values
3. Optional: Setting initial values
4. Initializing the simulation engine
5. Solving the ODE System (calculating the outputs)

Depending on the simulation of interest, some steps might take much longer than others. For example, a simulation with multiple administrations simulated over many months will load and initialize quickly (1 and 4) while calculating (5) will be the critical step. On the other hand, a simulation with dozens of compounds will also take time to load and initialize (1 and 4). 


## Running a single simulation once

This is by far the easiest use cases. A simulation is loaded from a pkml file and the outputs are calculated. In this use case, step 4 and 5 are automatically lumped together in one single call
See [Run Simulations](run-simulation.html) for more details. 


## Running a single simulation multiple times and varying some parameters/initial values between each run

This is already a use case where the optimal and most efficient way depends drastically on the simulation at hand.
Let's assume that we want to calculate the outputs of a simulation for different doses

A simple implementation could follow the following idea, where a simulation is loaded once, the dose value is updated and the simulation is calculated for each dose of interest. 
```{r SequentialRun, eval=TRUE}
library(ospsuite)
# Load and run the simulation
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)
doseParameter <- getAllParametersMatching(toPathString("Applications", "**", "Dose"), sim)[[1]]

# run for dose 100mg
doseParameter$value <- toBaseUnit(doseParameter, 100, "mg")
result100 <- runSimulations(simulations = sim)

# run for dose 200mg
doseParameter$value <- toBaseUnit(doseParameter, 200, "mg")
result200 <- runSimulations(simulations = sim)

# ...

# run for dose 500mg
doseParameter$value <- toBaseUnit(doseParameter, 500, "mg")
result500 <- runSimulations(simulations = sim)
```

This implementation is fairly easy to understand, but has potentially one big performance bottleneck. Each simulation run is performed **sequentially** (e.g. one at a time). That means that there will not be any performance boost from having multiple cores on the machine.
If the simulation only takes a few seconds to run, this solution is absolutely acceptable. However, if we assume that a simulation takes 1h to run, this code would take over 5h to complete, as each `runSimulations` call would be executed sequentially. 

When the simulation run time (5) is much greater than the loading time (1) and initialization time (4), a better approach is to load the simulation multiple times and run the simulation concurrently (e.g. in parallel)

Consider the following implementation:
```{r ParallelRun, eval=TRUE}
library(ospsuite)
# Load and run the simulation
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

loadSimulationWithDose <- function(doseInMg) {
  sim <- loadSimulation(simFilePath, loadFromCache = FALSE)
  doseParameter <- getAllParametersMatching(toPathString("Applications", "**", "Dose"), sim)[[1]]
  doseParameter$value <- toBaseUnit(doseParameter, doseInMg, "mg")
  return(sim)
}

# Creates 5 instances of a simulation (This is very fast for typical simulations)
sim100 <- loadSimulationWithDose(doseInMg = 100)
sim200 <- loadSimulationWithDose(doseInMg = 200)
sim300 <- loadSimulationWithDose(doseInMg = 300)
sim400 <- loadSimulationWithDose(doseInMg = 400)
sim500 <- loadSimulationWithDose(doseInMg = 500)


# Runs the simulation in parallel
results <- runSimulations(simulations = list(sim100, sim200, sim300, sim400, sim500))

# Results in now a list of SimulationResults
```

This implementation is also fairly straight forward. We load the simulation multiple times from the same file (note the usage of `loadFromCache=FALSE` to ensure that we always get a new instance of a simulation and not the same instance) and we set the dose. Then, we call the `runSimulations` function with the list of simulations. The engine will run these simulations in parallel, which means that the overall execution time should be slightly more than 1h, as each run would be executed **at the same time**.


## Running one or more simulations multiple times in a loop

This is a more advanced use case, typically when implementing some kind of optimization or sensitivity algorithms. The ospsuite-r package introduces the concept of a **simulation batch**. With a simulation batch, the variable parameters (or initial values) of a simulation to vary between each run are specified explicitely. This allows the simulation engine to speed up the system significantly as some equations can be rewritten and simplified. A simulation batch also keeps the simplified simulation in memory. That means that running a simulation again with a new set of values will be much faster, as the initialization phase (4) is only done if required and not for every run. 

Consider the following implementation:
```{r BatchRun, eval=TRUE}
library(ospsuite)
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

# We create two different instances of the same simulation to illustrate the usage.
sim1 <- loadSimulation(simFilePath, loadFromCache = FALSE)
sim2 <- loadSimulation(simFilePath, loadFromCache = FALSE)

# define the list of parameter that will be varied in the run. We will vary 2 parameters: Lipophilicity and Permeability 
parameterPaths <- c("Aciclovir|Lipophilicity", "Aciclovir|Permeability")

# define a first simulation batch
simBatch1 <- createSimulationBatch(simulation = sim1, parametersOrPaths = parameterPaths)

# For this new batch, we will vary Molecular Weight
simBatch2 <- createSimulationBatch(simulation = sim2, parametersOrPaths = c("Aciclovir|Molecular weight"))

simulationBatches <- list(simBatch1, simBatch2)
# now setting some parameter run values (the size of the array should match the number of parameters to vary for each batch
simBatch1$addRunValues(parameterValues = c(1, 2))
simBatch1$addRunValues(parameterValues = c(3, 4))
simBatch1$addRunValues(parameterValues = c(5, 6))

# We only have one parameter to vary for simBatch2, therefor only one value to set
simBatch2$addRunValues(parameterValues = 150)
simBatch2$addRunValues(parameterValues = 200)
simBatch2$addRunValues(parameterValues = 300)
simBatch2$addRunValues(parameterValues = 400)


# So far, we created 2 simulation batches, one with 3 parameter sets and the other one with 4. That means that 3 run slots will be allocated for simBatch1 and 4 will be allocated for simBatch2. These slots will remain in memory until the batch object is discarded

# Now we run the simulation batches. By doing so, 7 runs (3 for simBatch1 and 4 for simBatch2) will be executed in parallel. 
# Please see documentation of runSimulationBatches for more details on how to match parameter set to results.
results <- runSimulationBatches(simulationBatches)

# We can now set more values to the batches and run them again. 
# Notes: 
#  - We do not have to always set the same number of values at the same time
#  - Previous runValues are automaticaly cleared when runSimulationBatches is called. 
simBatch1$addRunValues(parameterValues = c(10, 20))
simBatch1$addRunValues(parameterValues = c(30, 40))
simBatch2$addRunValues(parameterValues = 500)
simBatch2$addRunValues(parameterValues = 200)

# this run will be much faster as the simulation won't be initialized again. Only the new value will be set as specified when adding new run values with addRunValues
results2 <- runSimulationBatches(simulationBatches)

# Once the simulation batches instances are not need anymore, they can be removed from the environment and the allocated memory cleared. This will be done automatically when the R session is terminated.
rm(simBatch1)
rm(simBatch2)
```

Usage of `SimulationBatch` is recommended for advanced scenarios where simulations expected to be run hundreds of times and where each second that can be spared will impact the performance significantly. 
