---
title: "Efficient calculation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Efficient calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

There are a few ways to run simulations using this package. 
The easiest way is, as expected, not necessarily the most efficient one. So depending on the use case, a different approach than what is outlined in [Run Simulations](run-simulation.html) might be more suited.

This vignette aims at describing some scenarios and the suggested methods to efficiently run simulations.

## Life cycle of a simulation run

In general, a user loads a simulation, updates some parameter values, calculates the simulation and retrieves the results. 
Here are some steps inherent to each simulation:

1. Loading a simulation
2. Optional: Setting parameter values
3. Optional: Setting initial values
4. Initializing the simulation engine
5. Solving the ODE System (calculating the outputs)

Depending on the simulation of interest, some steps might take much longer than others. For example, a simulation with multiple administrations simulated over many months will load and initialize quickly (1 and 4) while calculating (5) will be the critical step. On the other hand, a simulation with dozens of compounds will also take time to load and initialize (1 and 4). 

## Running a single simulation once

This is by far the easiest use cases. A simulation is loaded from a .pkml file and the outputs are calculated. In this use case, step 4 and 5 are automatically lumped together in one single call
(See [Run Simulations](run-simulation.html) for more details).

## Running a single simulation multiple times and varying some parameters/initial values between each run

This is already a use case where the optimal and most efficient way depends drastically on the simulation at hand.
Let's assume that we want to calculate the outputs of a simulation for different doses.

A simple implementation could follow the following idea, where a simulation is loaded once, the dose value is updated and the simulation is calculated for each dose of interest. 

```{r SequentialRun}
library(ospsuite)

# Load and run the simulation
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)
doseParameter <- getAllParametersMatching(toPathString("Applications", "**", "Dose"), sim)[[1]]

# run for dose 100mg
doseParameter$value <- toBaseUnit(doseParameter, 100, "mg")
result100 <- runSimulations(simulations = sim)

# run for dose 200mg
doseParameter$value <- toBaseUnit(doseParameter, 200, "mg")
result200 <- runSimulations(simulations = sim)

# ...

# run for dose 500mg
doseParameter$value <- toBaseUnit(doseParameter, 500, "mg")
result500 <- runSimulations(simulations = sim)
```

This implementation is fairly easy to understand, but has potentially one big performance bottleneck. Each simulation run is performed **sequentially** (i.e. one at a time). That means that there will not be any performance boost from having multiple cores on the machine.
If the simulation only takes a few seconds to run, this solution is absolutely acceptable. However, if we assume that a simulation takes 1h to run, this code would take over 5h to complete, as each `runSimulations` call would be executed sequentially. 

When the simulation run time (5) is much greater than the loading (1) and initialization time (4), a better approach is to load the simulation multiple times and run the simulation concurrently (i.e. in parallel).

Consider the following implementation:

```{r ParallelRun, eval=TRUE}
library(ospsuite)
# Load and run the simulation
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

loadSimulationWithDose <- function(doseInMg) {
  sim <- loadSimulation(simFilePath, loadFromCache = FALSE)
  doseParameter <- getAllParametersMatching(toPathString("Applications", "**", "Dose"), sim)[[1]]
  doseParameter$value <- toBaseUnit(doseParameter, doseInMg, "mg")
  return(sim)
}

# Creates 5 instances of a simulation (This is very fast for typical simulations)
sim100 <- loadSimulationWithDose(doseInMg = 100)
sim200 <- loadSimulationWithDose(doseInMg = 200)
sim300 <- loadSimulationWithDose(doseInMg = 300)
sim400 <- loadSimulationWithDose(doseInMg = 400)
sim500 <- loadSimulationWithDose(doseInMg = 500)


# Runs the simulation in parallel
results <- runSimulations(simulations = list(sim100, sim200, sim300, sim400, sim500))

# Results in now a list of SimulationResults
```

This implementation is also fairly straight forward. We load the simulation multiple times from the same file (note the usage of `loadFromCache = FALSE` to ensure that we always get a new instance of a simulation and not the same instance) and we set the dose. Then, we call the `runSimulations` function with the list of simulations. The engine will run these simulations in parallel, which means that the overall execution time should be slightly more than 1h, as each run would be executed **at the same time**.


## Optimized loading for running simulations with pre-defined set of variable parameters

This is a more advanced use case, typically when implementing some kind of optimization or sensitivity algorithms. The `{ospsuite}` package introduces the concept of a **simulation batch**. With a simulation batch, the variable parameters (or initial values) of a simulation to vary between each run are specified explicitly during the creation of the simulation batch. This allows the simulation engine to speed up the system significantly as some equations can be rewritten and simplified. A simulation batch also keeps the simplified simulation in memory. That means that running a simulation again with a new set of values will be much faster, as the initialization phase (4) is only done if required and not for every run. 

Consider the following implementation. We first create two simulation batches for the same model
and define the parameter values that can be changed between runs. Note that, as opposed to 
the `Simulation` object, we do not have access to any parameter/species values nor can we change
the outputs or simulation time.

```{r BatchRunInit, eval=TRUE}
library(ospsuite)
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")

# We load the simulation for which the batches will be created
sim1 <- loadSimulation(simFilePath, loadFromCache = FALSE)

# define the list of parameter that will be varied between the runs. 
# For the first batch, we will vary 2 parameters: Lipophilicity and Permeability 
parameterPaths <- c("Aciclovir|Lipophilicity", "Aciclovir|Permeability")

# define a first simulation batch
simBatch1 <- createSimulationBatch(simulation = sim1, parametersOrPaths = parameterPaths)

# for the second batch, we will vary Molecular Weight
simBatch2 <- createSimulationBatch(
  simulation = sim1, 
  parametersOrPaths = c("Aciclovir|Molecular weight")
  )

simulationBatches <- list(simBatch1, simBatch2)
```

In the next step, we define multiple runs of the simulation batch by adding different
parameter (and/or species start) values set. Each set of values correspond to one run. The benefit of this approach is that the initialization steps (1 and 4) are only performed once. After that, only the execution time (step 5) will have an impact on the total performance.

```{r BatchRunEnque, eval=TRUE}
# now setting some parameter run values (the size of the array should match 
# the number of parameters to vary for each batch
simBatch1$addRunValues(parameterValues = c(1, 2))
simBatch1$addRunValues(parameterValues = c(3, 4))
simBatch1$addRunValues(parameterValues = c(5, 6))

# We only have one parameter to vary for simBatch2, therefore only one value to set
simBatch2$addRunValues(parameterValues = 150)
simBatch2$addRunValues(parameterValues = 200)
simBatch2$addRunValues(parameterValues = 300)
simBatch2$addRunValues(parameterValues = 400)
```

So far, we created 2 simulation batches, one with 3 parameter sets and the 
other one with 4. That means that 3 runs will be enqueued for simBatch1 and 
4 will be enqueued for simBatch2. Each run gets a unique id that can later be used
to correctly assign simulation results to the simulated set of parameters.

```{r BatchRunRun, eval=TRUE}
# Now we run the simulation batches. 
# By doing so, 7 runs (3 for simBatch1 and 4 for simBatch2) will be executed in parallel. 
# Please see documentation of runSimulationBatches for more details.
# The resulting output is a named list, where the names are the ids of the enqueued runs.
results <- runSimulationBatches(simulationBatches)
print(names(unlist(results)))
```


The enqueued run values are cleared after calling `runSimulationBatches()`, so executing the run
again would result in an empty results list. We can now set more values to the batches and run them again. 
Notes: 
  - We do not have to always set the same number of values at the same time
  - Previous `runValues` are automatically cleared when `runSimulationBatches()` is called. 
  
```{r BatchRunRunNewValues, eval=TRUE}  
simBatch1$addRunValues(parameterValues = c(10, 20))
simBatch1$addRunValues(parameterValues = c(30, 40))
simBatch2$addRunValues(parameterValues = 500)
simBatch2$addRunValues(parameterValues = 200)

# this run will be much faster as the simulation won't be initialized again. 
# Only the new value will be set as specified when adding new run values with addRunValues
results2 <- runSimulationBatches(simulationBatches)

# Once the simulation batches instances are not needed anymore, they can be removed 
# from the environment and the allocated memory cleared. 
# This will be done automatically when the R session is terminated.
rm(simBatch1)
rm(simBatch2)
```

Usage of `SimulationBatch` is recommended for advanced scenarios where simulations
expected to be run hundreds of times and where each second that can be spared
will impact the performance significantly. 
