---
title: "Loading a simulation and accessing entities"
output: 
  rmarkdown::html_vignette:
    keep_md: yes
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Loading a simulation and accessing entities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Loading a simulation
In general, every workflow starts with loading a simulation by calling the `loadSimulation()` function. The function receives the full path to the PKML file (with extension *.pkml) exported from PK-Sim or MoBi and returns the corresponding simulation object.

```{r loadSim}
library(ospsuite)

simFilePath <- file.path(getwd(), "..", "tests", "data", "Aciclovir.pkml", fsep = .Platform$file.sep)

sim <- loadSimulation(simFilePath)
```

## Accessing entities of the model and their properties - the path concept

Once the simulation is loaded, it is possible to retrieve various entities of the model. The most important entities are **containers**, **molecules**, and **parameters**. The methods `getContainer`, `getMolecule`, and `getParameter` search for the respective entity with the given `path` located under a `container`. `path` is a string where the elements of the path (i.e., containers in the hierarchy of the simulation) are separated by `|`. `container` is an instance of the `Container`-class within the model structure the path is *relative* to. In most cases, `container` is the `Simulation` object created by calling `loadSimulation(pkmlSimulationFile)`.

```{r getEntities}
#Get the container "Liver"
livContainer <- getContainer("Organism|Liver", sim)
print(livContainer)

#Get the molecule Aciclovir located in kindey intracellular space
moleculeInKid <- getMolecule("Organism|Kidney|Intracellular|Aciclovir", sim)
print(moleculeInKid)

#Get the parameter volume of the liver interstitial space.
# Note that the path used is relative to the liver container
livParam <- getParameter("Interstitial|Volume", livContainer)
print(livParam)
```

The functions `getAllContainersMatching`, `getAllMoleculesMatching`, and `getAllParametersMatching` return a list of objects representing 
all entities whose paths match the generic paths provided in the list `paths` located under `container`. Generic paths are constructed by using the wildcard symbols `*` (exactly one occurrence of any element) or `**` (zero or more occurrences of any element).

```{r getAllEntitiesMatching}
# Get the parameter `Volume` of the intracellular space of all organs,
# with exactly one path element before `Intracellular`
volumeParams <- getAllParametersMatching("Organism|*|Intracellular|Volume", sim)
length(volumeParams)
#The PBPK model has 15 organs with an "Intracellular" sub-compartment

# Get the parameter `Volume` of the intracellular space of all organs,
# no matter how many sub-containers the organ has
volumeParams <- getAllParametersMatching("Organism|**|Intracellular|Volume", sim)
length(volumeParams)
# The list also includes parameters of organs like "Liver|Periportal",
# or the mucosal compartments of the intestine.
```
Note that the path `"Organism|Kidney|*|Intracellular|Volume"` will return no parameters in the standard models, as there are no sub-containers between `Kidney` and `Intracellular`. 
In contrast, `"Organism|Kidney|**|Intracellular|Volume"` is a valid path.

The functions `getAllXXXMatching` can also be used to retrieve entities from multiple paths:

```{r getAllEntitiesMatching_multiplePaths}
#Get the molecule Aciclovir located in `Liver|Periportal|Intracellular` and `VenousBlood|Plasma`
molecules <- getAllMoleculesMatching(c("Organism|Liver|Periportal|Intracellular|Aciclovir",
                                       "Organism|VenousBlood|Plasma|Aciclovir"), sim)
print(molecules)
```

The entities possess various properties that can be accessed through their objects. The most important properties for a container are:

```{r containerProperties}
# Path of the container
livContainer$path

# Parent container
livContainer$parentContainer
```

The most important properties for a molecule are:

```{r moleculeProperties}
#Initial value of the molecule
moleculeInKid$value

# Dimension of the molecule. See section "Unit conversion" for more information.
moleculeInKid$dimension

# Is the initial value defined by a formula?
moleculeInKid$isFormula

# Type of the formula. CONSTANT if the value is defined by a constant.
moleculeInKid$formula
```
The most important properties for a parameter are:

```{r parameterProperties}
#Initial value of the parameter
livParam$value

# Dimension of the parameter. See section "Unit conversion" for more information.
livParam$dimension

# Base unit of the parameter. See section "Unit conversion" for more information.
livParam$unit

# Is the initial value defined by a formula?
livParam$isFormula

# Type of the formula. CONSTANT if the value is defined by a constant.
livParam$formula
```

## Loading simulation tree
A convenient way to traverse the simulation structure is given by the method \code{getSimulationTree}. The method generates a tree-like list of all paths within the simulation. Each element of the tree contains all the sub-containers of the element. The final elements are strings representing the path to the entity.

```{r simTree}
# Load simulation
simFilePath <- file.path(getwd(), "..", "tests", "data", "Aciclovir.pkml", fsep = .Platform$file.sep)
sim <- loadSimulation(simFilePath)

# Create simulation tree
simTree <- getSimulationTree(sim)
 
# Calling simTree would list all entities within the simulation
# Accessing the parameter "Organism|Weight"
simTree$Organism$Weight
# 
# Getting all entities located under "Organism|Liver|Periportal|Intracellular"
entitiesList <- simTree$Organism$Liver$Periportal$Intracellular
```
