---
title: "Observed data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Observed data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`{ospsuite}` offers a concept of storing and processing numerical x-y data in a 
unified format by implementing a `DataSet` class. `DataSet` objects standardize 
handling of (observed) data coming from different sources, such as excel files 
using the data importer functionality of the OSPS, loaded from *.pkml, or manually 
created. This vignette gives an overview of the options to create `DataSet` objects
 and combine them into grouped data sets using the `DataCombined` class.

## DataSet

A `DataSet` object stores numerical data pairs - typically time 
as x values and measurement as y values - and optionally the measurement error . All values have a *dimension*  and a *unit* (see [Dimensions and Units](unit-conversion.html) 
for more information). Furthermore, each `DataSet` must have a *name*. When creating 
a `DataSet` from scratch (e.g. when the user wants to manually input observed data), 
a name must be provided:

```{r newDataSet}
library(ospsuite)

# Create an empty data set
dataSet <- DataSet$new("My data set")
```

After creation, the `DataSet` does not hold any data. The default dimension and 
unit for the x values is `Time` and `h`, respectively. The default dimension and 
unit for the y values is `Concentration (mass)` and `mg/l`, respectively. The 
dimension of the error values always corresponds to the dimension of the y values, 
though the units may differ.

Setting numerical values (or overwriting current values) is performed by the 
`$setValues()` method:


```{r setValues}
dataSet$setValues(xValues = c(1, 2, 3, 4),
                  yValues = c(0, 0.1, 0.6, 10),
                  yErrorValues = c(0.001, 0.001, 0.1, 1))

print(dataSet)
```

The user can change the dimensions and units of the values. After changing the 
dimension, the unit is automatically set to the *base unit*  of the dimension. 
Changing the dimension or unit *does not*  transform the values.


```{r setDimensionUnit}
# Print x, y, and error values
dataSet$xValues
dataSet$yValues
dataSet$yErrorValues

# Change the unit of x-values
dataSet$xUnit <- ospUnits$Time$min
# Print the x values - they did not change
dataSet$xValues

# Change dimension of y-values
dataSet$yDimension <- ospDimensions$Amount

print(dataSet)

# Change the units of y values and error values - they are now different!
dataSet$yUnit <- ospUnits$Amount$mol
dataSet$yErrorUnit <- ospUnits$Amount$pmol

print(dataSet)
```

Two types of error values are supported - arithmetic error (default) and geometric 
error, the latter being given in fraction. The user can change error type:

```{r setErrorType}
# Default error type is "ArithmeticStdDev"
dataSet$yErrorType

# Change error type to geometric
dataSet$yErrorType <- DataErrorType$GeometricStdDev
# Error unit is "Unitless" for dimension "Fraction".
dataSet$yErrorUnit

# Changing error type to arithmetic will set the dimension and unit of the error 
# to the same dimension and unit as the y values
dataSet$yErrorType <- DataErrorType$ArithmeticStdDev

print(dataSet)
```

A `DataSet` can store any kind of text meta data as meta data name - values pairs
 and can be added by the `addMetaData()` method:
 
```{r addMetaData}
# Add new meta data entries
dataSet$addMetaData(name = "Molecule",
                    value = "Aciclovir")
dataSet$addMetaData(name = "Organ",
                    value = "Muscle")

# Print meta data of the DataSet
print(dataSet$metaData)
```
 
A `DataSet` or multiple `DataSet`s  can be converted to `data.frame` (or `tibble`) 
to be processed in establised R workflows:

```{r dataSetToDataFrame}
# Create a second data set
dataSet2 <- DataSet$new(name = "Second data set")
dataSet2$setValues(xValues = c(1, 2, 3, 4, 5),
                   yValues = c(1, 0, 5, 8, 0.1))

# Convert data sets to a tibble
myTibble <- dataSetToTibble(dataSets = c(dataSet, dataSet2))

print(myTibble)
```

## Importing data

Creating `DataSet` objects from scratch is a rather advanced use case. Typically, 
observed data are loaded either from `*.pkml` files exported from PK-Sim or MoBi,
 or imported from Excel files. The function `loadDataSetFromPKML()` loads data from
 the `*.pkml` file. Complementary to this function is the function `saveDataSetToPKML()`
 that allows to export any `DataSet` to a `*.pkml` that can be loaded e.g. in MoBi.

```{r loadDataSetFromPKML}
# Load a data set from PKML
filePath <- system.file("extdata", "ObsDataAciclovir_1.pkml", package = "ospsuite")

dataSet <- loadDataSetFromPKML(filePath = filePath)

print(dataSet)
```

Another (and probably the most important) way to create `DataSet` objects is to 
import data from excel files. The function `loadDataSetsFromExcel()` utilizes the 
data import functionality implemented in PK-Sim and MoBi and returns a set of 
`DataSet` objects. For description of the supported file formats and configurations, 
please refer to the [OSPS documentation](https://docs.open-systems-pharmacology.org/shared-tools-and-example-workflows/import-edit-observed-data).

Loading observed data from an Excel sheet requires an `ImporterConfiguration`. 
The configuration describes mapping of excel sheet columns to numerical data (e.g. 
which column contains the x values) or meta data (e.g., description of the 
applied dose). One way to obtain such configuration is to create it in PK-Sim 
or MoBi, save it (as an `*.xml`) file, and load it in R with the `loadDataImporterConfiguration()` 
function:

```{r loadDataImporterConfiguration}
# Load a configuration from xml file
filePath <- system.file("extdata", "dataImporterConfiguration.xml", package = "ospsuite")
importerConfiguration <- loadDataImporterConfiguration(configurationFilePath = filePath)

print(importerConfiguration)
```

A data importer configuration can also be created from scratch and has to be 
manually populated by the user. Alternatively, the user can let the software 
"guess" the configuration for a given excel sheet:

```{r createDataImporterConfigurationFor}
# Excel file
excelFilePath <- system.file("extdata", "CompiledDataSet.xlsx", package = "ospsuite")
sheetName <- "TestSheet_1"

# Create importer configuration for the excel sheet
importerConfiguration_guessed <- createImporterConfigurationForFile(filePath =  excelFilePath,
                                                            sheet = sheetName)

print(importerConfiguration)
```

It is important to manually check the created configuration, as the automated 
configuration recognition cannot cover all possible cases.

If only specific sheets from the excel file should be imported, they can be 
specified in the `ImporterConfiguration`. The following example loads the sheets 
`TestSheet_1` and `TestSheet_1_withMW`:

```{r loadSheets}
# Excel file
excelFilePath <- system.file("extdata", "CompiledDataSet.xlsx", package = "ospsuite")
sheetName <- "TestSheet_1"

# Create importer configuration for the excel sheet
importerConfiguration_guessed <- createImporterConfigurationForFile(filePath =  excelFilePath,
                                                            sheet = sheetName)
# Add sheet names to the configuration
importerConfiguration_guessed$sheets <- c("TestSheet_1", "TestSheet_1_withMW")

# Load data
dataSets <- loadDataSetsFromExcel(xlsFilePath = excelFilePath,
                      importerConfigurationOrPath = importerConfiguration_guessed)

```

Currently, `DataImporterConfiguration` created from scratch or for a specific 
data sheet does not support all features of importer configuration, such as 
specifying the column containing the molecular weight of the measured molecule 
or the LLOQ values. It is therefore recommended to use importer configurations 
created in PK-Si or MoBi.
