---
title: "Calculating PK parameters of simulation outputs."
output: rmarkdown::html_vignette
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Calculating PK parameters of simulation outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

PK parameters such as AUC, C_max, etc., can be calculated for all outputs of a simulation. First, simulation results must be calculated, and the `SimulationResults` object is then passed to the method `calculatePKAnalyses`. For the list of calculated PK parameters and their description, please refer to [OSPS documentation](https://docs.open-systems-pharmacology.org/working-with-pk-sim/pk-sim-documentation/pk-sim-simulations#pk-analysis-view).


```{r calculatePKAnalyses}
library(ospsuite)

# Load  the simulation
simFilePath <- file.path(getwd(), "..", "tests", "data", "Aciclovir.pkml", fsep = .Platform$file.sep)
sim <- loadSimulation(simFilePath)

# Run the simulation
simulationResults <- runSimulation(simulation = sim)

# Calculate PK-analyses
pkAnalysis <- calculatePKAnalyses(results = simulationResults)

# Get the path of the simulated output
outputPath <- simulationResults$allQuantityPaths[[1]]
print(outputPath)

# Get all calculated PK parameters for the output path
allPkParams <- pkAnalysis$allPKParametersFor(outputPath)
print(allPkParams)

# Get C_max parameter
c_max <- pkAnalysis$pKParameterFor(quantityPath = outputPath, pkParameter = "C_max")
```

The function `calculatePKAnalyses` returns an object of the class `SimulationPKAnalyses`, which allows to retrieve either a certain PK parameter for an output path, or all calculated PK parameters for an output path.

The methods `$allPKParametersFor` and `$pKParameterFor` return a object (or a list of objects) of the class `QuantityPKParameter`, with the fields `$name` which is the name of the pk-parameter (e.g. "C_max"), `$quantityPath` the path of the output the parameter has been calculated for, and `$values` the value (or list of values for a population simulation).

```{r QuantityPKParameter}
# Get C_max parameter
c_max <- pkAnalysis$pKParameterFor(quantityPath = outputPath, pkParameter = "C_max")

print(c_max)

c_max$name
c_max$quantityPath
c_max$values
```

In case of a population simulation, `$values` return a list of values calculated for each individual.

## Import and export of PK-analyses
Population analysis calculated in R can be exported to a *.csv file and loaded in PK-Sim, and vice versa.

```{r PKAnalysesExport}
# Load and run the simulation
simFilePath <- file.path(getwd(), "..", "tests", "data", "Aciclovir.pkml", fsep = .Platform$file.sep)
sim <- loadSimulation(simFilePath)
simulationResults <- runSimulation(simulation = sim)

# Calculate PK-analysis
pkAnalysis <- calculatePKAnalyses(results = simulationResults)

# Export to csv
csvPKAnalysisPath <- file.path(getwd(), "..", "tests", "data", "PKAnalyses.csv", fsep = .Platform$file.sep)
exportPKAnalysesToCSV(pkAnalyses = pkAnalysis, filePath = csvPKAnalysisPath)

# Load from csv
pkAnalysisLoaded <- importPKAnalysesFromCSV(filePath = csvPKAnalysisPath, simulation = sim)
```
