---
title: "Plotting with ospsuite.plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting with ospsuite.plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  out.width = "100%",
  fig.height = 6,
  fig.width = 8,
  fig.showtext = TRUE
)
```

## Introduction

The `{ospsuite}` package provides a set of plotting functions based on the `{ospsuite.plots}` library. These functions are designed to work seamlessly with `DataCombined` objects and provide advanced visualization capabilities for pharmacometric data analysis.

This vignette describes the exported plotting functions from the `plot-with-ospsuite-plots.R` file:

- `plotTimeProfile()` - Creates time profile plots
- `plotPredictedVsObserved()` - Creates predicted vs observed scatter plots
- `plotResidualsVsCovariate()` - Creates residual plots against time, observed, or predicted values
- `plotResidualsAsHistogram()` - Creates histogram plots of residuals
- `plotQuantileQuantilePlot()` - Creates Q-Q plots for assessing residual distribution

All these functions return `ggplot2` objects that can be further customized and saved.

For more information on the underlying library:
- `{ospsuite.plots}`: [GitHub repository](https://github.com/Open-Systems-Pharmacology/OSPSuite.Plots)


## Initial Setup

Before creating plots with `ospsuite.plots`, it's important to initialize the plotting environment properly. This involves two key steps:

1. **Set default plotting options** using `ospsuite.plots::setDefaults()`
2. **Configure the watermark option** using `options(ospsuite.plots.watermark_enabled = TRUE)` or `options(ospsuite.plots.watermark_enabled = FALSE)` 

```{r setup-options}
# library(ospsuite)
devtools::load_all()

# Set default plotting options for ospsuite.plots
# Note: ospsuite.plots is imported by ospsuite, so we use :: for clarity
ospsuite.plots::setDefaults()

# Enable watermark for plots (optional)
options(ospsuite.plots.watermark_enabled = TRUE)
```

The `setDefaults()` function initializes various plotting defaults that ensure consistent appearance across all plots. The watermark option allows you to add a watermark to your plots.  
Refer to the `{ospsuite.plots}` package documentation for details.

## Setting up the data

Next, let's create a `DataCombined` object that we will use throughout this vignette:

```{r setup}
# Load simulation
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)
simResults <- runSimulations(sim)[[1]]
outputPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"

# Load observed data
obsData <- lapply(
  c(
    "ObsDataAciclovir_1.pkml",
    "ObsDataAciclovir_2.pkml",
    "ObsDataAciclovir_3.pkml"
  ),
  function(x) {
    loadDataSetFromPKML(system.file("extdata", x, package = "ospsuite"))
  }
)
names(obsData) <- lapply(obsData, function(x) x$name)

# Create DataCombined object
myDataCombined <- DataCombined$new()

myDataCombined$addSimulationResults(
  simulationResults = simResults,
  quantitiesOrPaths = outputPath,
  groups = "Aciclovir PVB"
)

myDataCombined$addDataSets(
  obsData$`Vergin 1995.Iv`,
  groups = "Aciclovir PVB"
)
```

### Population Simulation Data

For demonstrating population-specific features, let's also create a `DataCombined` object with population simulation results:

```{r setup-population}
# Load population from CSV file
popFilePath <- system.file("extdata", "pop.csv", package = "ospsuite")
population <- loadPopulation(csvPopulationFile = popFilePath)

# Run population simulation
popResults <- runSimulations(simulations = sim, population = population)[[1]]

# Create DataCombined object for population
myPopDataCombined <- DataCombined$new()

myPopDataCombined$addSimulationResults(
  simulationResults = popResults,
  quantitiesOrPaths = outputPath,
  groups = "Aciclovir Population"
)

myPopDataCombined$addDataSets(
  obsData$`Vergin 1995.Iv`,
  groups = "Aciclovir Population"
)
```

## Time Profile Plots

The `plotTimeProfile()` function creates time profile plots showing observed and simulated data over time. This is one of the most common visualizations in pharmacometric analysis.

```{r timeprofile}
plotTimeProfile(myDataCombined)
```

### Population Data Aggregation

When working with population simulations, you can control how the data is aggregated using the `aggregation` parameter. Options include:

- `"quantiles"` (default) - Shows median and specified quantiles
- `"arithmetic"` - Shows arithmetic mean with standard deviations
- `"geometric"` - Shows geometric mean with standard deviations

Here's an example using the population simulation data:

```{r timeprofile-population}
# Plot population data with default quantile aggregation
plotTimeProfile(myPopDataCombined, yScale = 'log')
```

You can customize the quantiles:

```{r timeprofile-aggregation}
plotTimeProfile(
  myPopDataCombined,
  yScale = 'log',
  aggregation = "quantiles",
  quantiles = c(0.1, 0.5, 0.9)
)
```

Alternatively, you can use arithmetic mean aggregation with standard deviation bands:

```{r timeprofile-arithmetic}
plotTimeProfile(
  myPopDataCombined,
  yScale = 'log',
  aggregation = "arithmetic",
  nsd = 1
)
```

### Custom Aesthetic Mappings

By default, the legend is created using the `group` variable of the `DataCombined` object.
You can customize the plot appearance using `ggplot2` aesthetic mappings. 

```{r timeprofile-mapping}
# Customize aesthetics
plotTimeProfile(
  myDataCombined,
  mapping = ggplot2::aes(linetype = dataType)
)
```

## Predicted vs Observed Plots

Beyond time profiles, you can assess model performance using goodness-of-fit plots. The `plotPredictedVsObserved()` function creates scatter plots comparing predicted (simulated) values to observed values. This helps assess model performance and identify systematic biases.

```{r predvsobs}
plotPredictedVsObserved(myDataCombined)
```

By default, the plot shows:
- Identity line (perfect agreement)
- Fold-distance lines (default: 2-fold range)

### Customizing Fold Distance

You can customize the fold-distance lines using the `comparisonLineVector` parameter:

```{r predvsobs-fold}
# Show 1.5-fold and 3-fold ranges
plotPredictedVsObserved(
  myDataCombined,
  comparisonLineVector = ospsuite.plots::getFoldDistanceList(folds = c(1.5, 3))
)
```

### Swapping Axes

By default, predicted values are on the y-axis and observed on the x-axis. You can swap these using the `predictedAxis` parameter:

```{r predvsobs-swap}
# Put predicted on x-axis, observed on y-axis
plotPredictedVsObserved(myDataCombined, predictedAxis = "x")
```

### Scaling Options

The `xyScale` parameter controls the axis scaling:

```{r predvsobs-scale}
# Use linear scale instead of log
plotPredictedVsObserved(myDataCombined, xyScale = "linear")
```

## Residuals vs Covariate Plots

The `plotResidualsVsCovariate()` function creates residual plots to assess systematic bias in the model. You can plot residuals against time, observed values, or predicted values.

### Residuals vs Observed Values

```{r resvsobs}
plotResidualsVsCovariate(myDataCombined, xAxis = "observed")
```

### Residuals vs Time

```{r resvstime}
plotResidualsVsCovariate(myDataCombined, xAxis = "time")
```

### Residuals vs Predicted Values

```{r resvspred}
plotResidualsVsCovariate(myDataCombined, xAxis = "predicted")
```

### Residual Scale Options

The `residualScale` parameter controls how residuals are calculated and displayed (this same parameter is used in `plotResidualsAsHistogram` and `plotQuantileQuantilePlot`):

- `"log"` (default) - Logarithmic residuals: `log(observed/predicted)`
- `"linear"` - Linear residuals: `observed - predicted`
- `"ratio"` - Ratio: `observed/predicted`

```{r resvsobs-scale}
# Use linear residuals
plotResidualsVsCovariate(
  myDataCombined,
  xAxis = "observed",
  residualScale = "linear"
)
```

## Residuals as Histogram

The `plotResidualsAsHistogram()` function creates a histogram of residuals, which helps assess the distribution of errors.

```{r reshist}
plotResidualsAsHistogram(myDataCombined)
```

By default, a normal distribution overlay is added to help assess normality. You can control this using the `distribution` parameter:

```{r reshist-dist, eval=FALSE}
# Without distribution overlay
plotResidualsAsHistogram(myDataCombined, distribution = 'none')
```

The `residualScale` parameter works the same as in `plotResidualsVsCovariate()`:

```{r reshist-scale, eval=FALSE}
# Linear residuals histogram
plotResidualsAsHistogram(myDataCombined, residualScale = "linear")
```

## Quantile-Quantile (Q-Q) Plot

The `plotQuantileQuantilePlot()` function creates a Q-Q plot to assess whether residuals follow a normal distribution.

```{r qqplot}
plotQuantileQuantilePlot(myDataCombined)
```

Points falling along the diagonal line indicate that the residuals follow a normal distribution. Deviations suggest non-normality.

```{r qqplot-scale, eval=FALSE}
# Use linear residuals
plotQuantileQuantilePlot(myDataCombined, residualScale = "linear")
```

## Automatic Unit Conversion

A key feature of these plotting functions is automatic unit conversion. For manual unit conversion of `DataCombined` objects outside of plotting, you can use the `convertUnits()` function. When using a `DataCombined` object with mixed units in plotting functions:

- The target unit is automatically determined by the most frequently occurring unit in the observed data
- If no observed data exists, the most common unit in simulated data is used
- Concentration dimensions (`Concentration (mass)` and `Concentration (molar)`) are treated as compatible
- Conversion between mass and molar concentrations is possible if molecular weight is available

This ensures that all data is displayed in consistent units without manual conversion.

## Handling Mixed Error Types

These functions automatically handle datasets with different error type specifications:

- If all data uses the same error type (`ArithmeticStdDev` or `GeometricStdDev`), it is used directly
- If data contains **mixed error types**, they are automatically converted to `yMin`/`yMax` bounds:
  - `ArithmeticStdDev`: `yMin = yValues - yErrorValues`, `yMax = yValues + yErrorValues`
  - `GeometricStdDev`: `yMin = yValues / yErrorValues`, `yMax = yValues * yErrorValues`

## Using data.table Instead of DataCombined

While these functions are designed to work with `DataCombined` objects, you can also provide a `data.table` directly. Use `toDataFrame()` to convert a `DataCombined` object to a data frame if needed. The table must include the following columns:

- `xValues`: Numeric time points or x-axis values
- `yValues`: Observed or simulated values (numeric)
- `group`: Grouping variable (factor or character)
- `name`: Name for the dataset (factor or character)
- `xUnit`: Unit of the x-axis values (character)
- `yUnit`: Unit of the y-axis values (character)
- `dataType`: Specifies data type—either `"observed"` or `"simulated"`

Optional columns:
- `yErrorType`: Type of y error (see `ospsuite::DataErrorType`)
- `yErrorValues`: Numeric error values
- `yMin`, `yMax`: Custom ranges for y-axis
- `IndividualId`: Used for aggregation of simulated population data
- `predicted`: Predicted values (required for residual plots)

## Further Customization

All plotting functions accept additional arguments that are passed to the underlying `ospsuite.plots` functions. This allows for extensive customization. Refer to the `{ospsuite.plots}` package documentation for details.

Additionally, since all functions return `ggplot2` objects, you can further modify them using standard `ggplot2` functions:

```{r custom, eval=FALSE}
library(ggplot2)

# Create a plot and customize it
p <- plotTimeProfile(myDataCombined)

# Add customizations
p <- p +
  theme_minimal() +
  labs(title = "My Custom Title") +
  theme(legend.position = "bottom")

print(p)
```

## Saving Plots

Since all functions return `ggplot2` objects, you can save them using `ospsuite.plots::exportPlot()`:

```{r save, eval=FALSE}
# Create a plot
myPlot <- plotTimeProfile(myDataCombined)

# Save to file using ospsuite.plots::exportPlot
ospsuite.plots::exportPlot(
  plotObject = myPlot,
  filePath = "timeprofile.png",
  width = 8,
  height = NULL,
  dpi = 300
)
```

This function is a wrapper around `ggsave()` that automatically adjusts the plot height based on content when `height = NULL`.

Alternatively, you can use directly the standard `ggsave()` function from `ggplot2`:

```{r save-ggsave, eval=FALSE}
# Save using ggsave
ggsave("timeprofile.png", myPlot, width = 8, height = 6, dpi = 300)
```




