---
title: "Plotting with ospsuite.plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting with ospsuite.plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  out.width = "100%",
  fig.height = 6,
  fig.width = 8,
  fig.showtext = TRUE
)
```

## Introduction

The `{ospsuite}` package provides a set of plotting functions based on the `{ospsuite.plots}` library. These functions are designed to work seamlessly with `DataCombined` objects and provide advanced visualization capabilities for pharmacometric data analysis.

This vignette describes the exported plotting functions from the `plot-with-ospsuite-plots.R` file:

- `plotTimeProfile()` - Creates time profile plots
- `plotPredictedVsObserved()` - Creates predicted vs observed scatter plots
- `plotResidualsVsCovariate()` - Creates residual plots against time, observed, or predicted values
- `plotResidualsAsHistogram()` - Creates histogram plots of residuals
- `plotQuantileQuantilePlot()` - Creates Q-Q plots for assessing residual distribution

All these functions return `ggplot2` objects that can be further customized and saved.

## Initial Setup

Before creating plots with `ospsuite.plots`, it's important to initialize the plotting environment properly. This involves two key steps:

1. **Set default plotting options** using `ospsuite.plots::setDefaults()`
2. **Configure the watermark option** using `options(ospsuite.plots.watermark_enabled = TRUE)`

```{r setup-options}
library(ospsuite)

# Set default plotting options for ospsuite.plots
# Note: ospsuite.plots is imported by ospsuite, so we use :: for clarity
ospsuite.plots::setDefaults()

# Enable watermark for plots (optional)
options(ospsuite.plots.watermark_enabled = TRUE)
```

The `setDefaults()` function initializes various plotting defaults that ensure consistent appearance across all plots. The watermark option allows you to add a watermark to your plots for publication or presentation purposes.

## Setting up the data

Next, let's create a `DataCombined` object that we will use throughout this vignette:

```{r setup}
# Load simulation
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)
simResults <- runSimulations(sim)[[1]]
outputPath <- "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)"

# Load observed data
obsData <- lapply(
  c("ObsDataAciclovir_1.pkml", "ObsDataAciclovir_2.pkml", "ObsDataAciclovir_3.pkml"),
  function(x) loadDataSetFromPKML(system.file("extdata", x, package = "ospsuite"))
)
names(obsData) <- lapply(obsData, function(x) x$name)

# Create DataCombined object
myDataCombined <- DataCombined$new()

myDataCombined$addSimulationResults(
  simulationResults = simResults,
  quantitiesOrPaths = outputPath,
  groups = "Aciclovir PVB"
)

myDataCombined$addDataSets(
  obsData$`Vergin 1995.Iv`,
  groups = "Aciclovir PVB"
)
```

## Time Profile Plots

The `plotTimeProfile()` function creates time profile plots showing observed and simulated data over time. This is one of the most common visualizations in pharmacometric analysis.

```{r timeprofile}
plotTimeProfile(myDataCombined)
```

### Population Data Aggregation

When working with population simulations, you can control how the data is aggregated using the `aggregation` parameter. Options include:

- `"quantiles"` (default) - Shows median and specified quantiles
- `"arithmetic"` - Shows arithmetic mean with standard deviations
- `"geometric"` - Shows geometric mean with standard deviations

```{r timeprofile-aggregation, eval=FALSE}
# For population data, you can specify quantiles
plotTimeProfile(myDataCombined, aggregation = "quantiles", quantiles = c(0.05, 0.5, 0.95))

# Or use arithmetic mean with standard deviations
plotTimeProfile(myDataCombined, aggregation = "arithmetic", nsd = 1)
```

### Custom Aesthetic Mappings

You can customize the plot appearance using `ggplot2` aesthetic mappings:

```{r timeprofile-mapping, eval=FALSE}
# Customize aesthetics
plotTimeProfile(
  myDataCombined,
  mapping = ggplot2::aes(color = group, linetype = name)
)
```

## Predicted vs Observed Plots

The `plotPredictedVsObserved()` function creates scatter plots comparing predicted (simulated) values to observed values. This helps assess model performance and identify systematic biases.

```{r predvsobs}
plotPredictedVsObserved(myDataCombined)
```

By default, the plot shows:
- Identity line (perfect agreement)
- Fold-distance lines (default: 2-fold range)

### Customizing Fold Distance

You can customize the fold-distance lines using the `comparisonLineVector` parameter:

```{r predvsobs-fold}
# Show 1.5-fold and 3-fold ranges
plotPredictedVsObserved(
  myDataCombined,
  comparisonLineVector = ospsuite.plots::getFoldDistanceList(folds = c(1.5, 3))
)
```

### Swapping Axes

By default, predicted values are on the y-axis and observed on the x-axis. You can swap these using the `predictedAxis` parameter:

```{r predvsobs-swap}
# Put predicted on x-axis, observed on y-axis
plotPredictedVsObserved(myDataCombined, predictedAxis = "x")
```

### Scaling Options

The `xyScale` parameter controls the axis scaling:

```{r predvsobs-scale, eval=FALSE}
# Use linear scale instead of log
plotPredictedVsObserved(myDataCombined, xyScale = "linear")
```

## Residuals vs Covariate Plots

The `plotResidualsVsCovariate()` function creates residual plots to assess systematic bias in the model. You can plot residuals against time, observed values, or predicted values.

### Residuals vs Observed Values

```{r resvsobs}
plotResidualsVsCovariate(myDataCombined, xAxis = "observed")
```

### Residuals vs Time

```{r resvstime}
plotResidualsVsCovariate(myDataCombined, xAxis = "time")
```

### Residuals vs Predicted Values

```{r resvspred}
plotResidualsVsCovariate(myDataCombined, xAxis = "predicted")
```

### Residual Scale Options

The `residualScale` parameter controls how residuals are calculated and displayed:

- `"log"` (default) - Logarithmic residuals: `log(observed/predicted)`
- `"linear"` - Linear residuals: `observed - predicted`
- `"ratio"` - Ratio: `observed/predicted`

```{r resvsobs-scale}
# Use linear residuals
plotResidualsVsCovariate(myDataCombined, xAxis = "observed", residualScale = "linear")
```

## Residuals as Histogram

The `plotResidualsAsHistogram()` function creates a histogram of residuals, which helps assess the distribution of errors.

```{r reshist}
plotResidualsAsHistogram(myDataCombined)
```

By default, a normal distribution overlay is added. You can control this using the `distribution` parameter:

```{r reshist-dist, eval=FALSE}
# Without distribution overlay
plotResidualsAsHistogram(myDataCombined, distribution = NULL)
```

The `residualScale` parameter works the same as in `plotResidualsVsCovariate()`:

```{r reshist-scale, eval=FALSE}
# Linear residuals histogram
plotResidualsAsHistogram(myDataCombined, residualScale = "linear")
```

## Quantile-Quantile (Q-Q) Plot

The `plotQuantileQuantilePlot()` function creates a Q-Q plot to assess whether residuals follow a normal distribution.

```{r qqplot}
plotQuantileQuantilePlot(myDataCombined)
```

Points falling along the diagonal line indicate that the residuals follow a normal distribution. Deviations suggest non-normality.

```{r qqplot-scale, eval=FALSE}
# Use linear residuals
plotQuantileQuantilePlot(myDataCombined, residualScale = "linear")
```

## Automatic Unit Conversion

A key feature of these plotting functions is automatic unit conversion. When using a `DataCombined` object with mixed units:

- The target unit is automatically determined by the most frequently occurring unit in the observed data
- If no observed data exists, the most common unit in simulated data is used
- Concentration dimensions (`Concentration (mass)` and `Concentration (molar)`) are treated as compatible
- Conversion between mass and molar concentrations is possible if molecular weight is available

This ensures that all data is displayed in consistent units without manual conversion.

## Handling Mixed Error Types

These functions automatically handle datasets with different error type specifications:

- If all data uses the same error type (`ArithmeticStdDev` or `GeometricStdDev`), it is used directly
- If data contains **mixed error types**, they are automatically converted to `yMin`/`yMax` bounds:
  - `ArithmeticStdDev`: `yMin = yValues - yErrorValues`, `yMax = yValues + yErrorValues`
  - `GeometricStdDev`: `yMin = yValues / yErrorValues`, `yMax = yValues * yErrorValues`

## Using data.table Instead of DataCombined

While these functions are designed to work with `DataCombined` objects, you can also provide a `data.table` directly. The table must include the following columns:

- `xValues`: Numeric time points or x-axis values
- `yValues`: Observed or simulated values (numeric)
- `group`: Grouping variable (factor or character)
- `name`: Name for the dataset (factor or character)
- `xUnit`: Unit of the x-axis values (character)
- `yUnit`: Unit of the y-axis values (character)
- `dataType`: Specifies data type—either `"observed"` or `"simulated"`

Optional columns:
- `yErrorType`: Type of y error (see `ospsuite::DataErrorType`)
- `yErrorValues`: Numeric error values
- `yMin`, `yMax`: Custom ranges for y-axis
- `IndividualId`: Used for aggregation of simulated population data
- `predicted`: Predicted values (required for residual plots)

## Further Customization

All plotting functions accept additional arguments that are passed to the underlying `ospsuite.plots` functions. This allows for extensive customization. Refer to the `{ospsuite.plots}` package documentation for details.

Additionally, since all functions return `ggplot2` objects, you can further modify them using standard `ggplot2` functions:

```{r custom, eval=FALSE}
library(ggplot2)

# Create a plot and customize it
p <- plotTimeProfile(myDataCombined)

# Add customizations
p <- p +
  theme_minimal() +
  labs(title = "My Custom Title") +
  theme(legend.position = "bottom")

print(p)
```

## Saving Plots

Since all functions return `ggplot2` objects, you can save them using `ggsave()`:

```{r save, eval=FALSE}
# Create a plot
myPlot <- plotTimeProfile(myDataCombined)

# Save to file
ggsave("timeprofile.png", myPlot, width = 8, height = 6, dpi = 300)
```

## Implementation Details

All plotting functions in this vignette make use of the `{ospsuite.plots}` library, which builds on the `{tlf}` (TLF Library) for creating high-quality pharmacometric visualizations. 

For more information:
- `{ospsuite.plots}`: [GitHub repository](https://github.com/Open-Systems-Pharmacology/OSPSuite.Plots)
- `{tlf}`: [Website](https://www.open-systems-pharmacology.org/TLF-Library/)

## Comparison with Other Plotting Functions

The `{ospsuite}` package also provides alternative plotting functions such as `plotIndividualTimeProfile()`, `plotObservedVsSimulated()`, etc. These are documented in the [Visualizations with DataCombined](data-combined-plotting.html) vignette.

The functions described in this vignette (`plotTimeProfile()`, `plotPredictedVsObserved()`, etc.) are newer implementations that leverage the `{ospsuite.plots}` library and offer enhanced features such as:

- Automatic unit conversion
- More flexible residual calculations
- Better handling of mixed error types
- Enhanced customization options

Choose the set of functions that best fits your workflow and requirements.
