---
title: "Running a simulation"
output: rmarkdown::html_vignette
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Running a simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Running individual simulation and retrieving the results
Once the simulation is loaded (see [Loading a simulation and accessing entities](load-get.html)), it can be run to produce an object of the class `SimulationResults`.

```{r runSim}
library(ospsuite)

# Load the simulation
simFilePath <- file.path(getwd(), "..", "tests", "data", "Aciclovir.pkml", fsep = .Platform$file.sep)
sim <- loadSimulation(simFilePath)

simulationResults <- runSimulation(simulation = sim)
print(simulationResults)
```

The advantage of storing the results in a object is the option to keep different resutls of the same simulation produced with different settings (e.g., model parameters).

Simulated time-value pairs for a specific ouput from the `SimulationResults`-object can be accessed with the method `getOutputValues`. The user can provide either the path(s) of the output (which can be a molecule, a parameter, or an observer), or the object(s) of the type `Molecule`, `Parameter`, or `Quantity` (for observers) with the argument `quantitiesOrPaths`. If no output is specified, all outputs available in the simulation results are returned.

The paths of all available outputs can be accessed via

```{r getAllOutputSelections}
simulationResults$allQuantityPaths
```

`getOutputValues` returns a list with two entries: `data` and `metadata`:

 * `data` is a dataframe with two predefined columns (IndividualId and Time) as well as one column for each requested output
 
     * `IndividualId` (not relevant for an individual simulation)
     * `Time` a vector with simulated time values (in minutes, equal for all outputs) 
     *  a vector with simulated entries for each output requested.

  * `metaData` is a list containing one entry for each requested output. Each entry contains information pertinent to the output such as its dimension or the unit.
  
  
```{r getOutputValues}
# Get simulated results by path
resultsPath <- simulationResults$allQuantityPaths[[1]]
print(resultsPath)

resultsData <- getOutputValues(simulationResults, quantitiesOrPaths = resultsPath)

resultsTime <- resultsData$data$Time
resultsValues <- resultsData$data$`Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)`

plot(resultsTime, resultsValues, type = "l")
```

The results can be stored in and imported from a *.csv file with the methods `exportResultsToCSV` and `importResultsFromCSV`.

## Adding new outputs
By default, only outputs that were selected in PK-Sim or MoBi prior to the export of the simulation to `pkml` are generated. The user can add new outputs to the simulation with the method `addOutputs`. The outputs can be provided either as objects of the type(s) `Molecule`, `Parameter`, or `Quantity`, or as path strings. The ouput list is a property of the `simulation`.
After adding or removing outputs, the corresponding simulation needs to be re-run in order to generate updated results.

```{r addOutputs}
# Clear the list of generated outputs
clearOutputs(sim)

# Add new outputs as objects
molecule <- getMolecule("Organism|Kidney|Intracellular|Aciclovir", sim)
observer <- getQuantity("Organism|Lumen|Aciclovir|Fraction dissolved", sim)

addOutputs(c(molecule, observer), simulation = sim)

# Add new ouputs as path strings
addOutputs(c("Organism|Lumen|Stomach|Aciclovir", 
             "Organism|PeripheralVenousBlood|Aciclovir|Whole Blood (Peripheral Venous Blood)"),
           simulation = sim)

# Run simulation
simulationResults <- runSimulation(simulation = sim)

# Retrieve all generated outputs (e.g. omitting the quantitiesOrPaths property
# will return all available values)
resultsData <- getOutputValues(simulationResults)

# Note that "Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)" 
# is not in the list of generated results any more
names(resultsData$data)
```

## Changing simulation intervals
The simulation interval (i.e., the simulation times at which results are stored) can are stored as the property `outputSchema` of a `Simulation` object.

```{r $outputSchema}
print(sim$outputSchema)
```

To change the simulation interval, the user can use one of the functions `clearOutputIntervals`, `addOutputInterval`, and `setOutputInterval`.

```{r changeOutputInterval, error = TRUE, purl = FALSE}
# Remove all output intervals - simulation not possible!
clearOutputIntervals(simulation = sim)
runSimulation(simulation = sim)

# Add an interval
addOutputInterval(simulation = sim, startTime = 0, endTime = 20, resolution = 60, intervalName = "highRes")
print(sim$outputSchema)

# Add a second interval
addOutputInterval(simulation = sim, startTime = 30, endTime = 2000, resolution = 4, intervalName = "lowRes")
print(sim$outputSchema)

# Replace the existing interval(s) with a new one
setOutputInterval(simulation = sim, startTime = 0, endTime = 2000, resolution = 4)
print(sim$outputSchema)
```
