---
title: "Sensitivity analysis"
output: rmarkdown::html_vignette
#output: pdf_document
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Population simulations
Population simulations can be easily performed in R by combining the simulation loaded from a *.pkml file with the population information created in PK-Sim and exported to CSV format (for details, please refer to [OSPS online documentation](https://docs.open-systems-pharmacology.org/working-with-pk-sim/pk-sim-documentation/pk-sim-creating-populations)) or created directly in R (see [Creating simulations](#SectionCreatePopulation)).

  ### Loading population file
  The method `loadPopulation` creates an object of the `Population` class that can be passed to the `runSimulation` method (see [Running simulations and retrieving the results](run-simulation.html)).

  ```{r loadPop}
  library(ospsuite)
  # Load population information from csv
  popFilePath <- file.path(getwd(), "..", "tests", "data", "pop_10.csv", fsep = .Platform$file.sep)
  myPopulation <- loadPopulation(csvPopulationFile = popFilePath)
  print(myPopulation)
  ```

  ## Creating populations {#SectionCreatePopulation}
  Similar to creating individual parameter sets (see [Creating individuals](create-individual.html)), a population is created from *population characteristics* created by calling the method `createPopulationCharacteristics`. To see the list of available values for the arguments `species` and `population` (only for human), use the enums `Species` and `HumanPopulation`, respectively. The returned object of type `PopulationCharacteristics` is then passed to the function `createPopulation` to generate a set of parameter values. The algoritghm behind is the same used in PK-Sim when creating an population.

  ```{r createPop}
  library(ospsuite)

  populationCharacteristics <- createPopulationCharacteristics(
    species = Species$Human,
    population = HumanPopulation$Asian_Tanaka_1996,
    numberOfIndividuals = 50,
    proportionOfFemales = 50,
    weightMin = 30,
    weightMax = 98,
    weightUnit = "kg",
    heightMin = NULL,
    heightMax = NULL,
    ageMin = 0,
    ageMax = 80,
    ageUnit = "year(s)")
  print(populationCharacteristics)

  # Create population from population characteristics
  myPopulation <- createPopulation(populationCharacteristics = populationCharacteristics)
  print(myPopulation)
  ```

  ## Running population simulation
  To run a population simulation, the `Population` object created by the `createPopulation` method must be passed to the `runSimulation` method:

    ```{r runPop}
  library(ospsuite)

  # Load simulation
  simFilePath <- file.path(getwd(), "..", "tests", "data", "Aciclovir.pkml", fsep = .Platform$file.sep)
  sim <- loadSimulation(simFilePath)

  # Run population simulation
  simulationResults <- runSimulation(simulation = sim, population = myPopulation)
  print(simulationResults)
  ```

  Population simulations are run in parallel on multi-core machines - one core simulates a subset of all individuals defined in the population. By default, the number of cores used equals the maximal number of logical cores available minus one.
  The user can change the default behavior by providing custom `SimulationRunOptions`.

  ```{r simulationRunOptions}
  # Load simulation
  simFilePath <- file.path(getwd(), "..", "tests", "data", "Aciclovir.pkml", fsep = .Platform$file.sep)
  sim <- loadSimulation(simFilePath)

  # Create a SimulationRunOptions object
  simRunOptions <- SimulationRunOptions$new()
  print(simRunOptions)

  # Change the maximal number of cores to use and show a progress bar during simulation
  simRunOptions$numberOfCores <- 3
  simRunOptions$showProgress <- TRUE

  # Run population simulation with cutom options
  populationResults <- runSimulation(simulation = sim, population = myPopulation, simulationRunOptions = simRunOptions)
  print(populationResults)
  ```

  Simulated time-value pairs for a specific ouput from the `SimulationResults`-object returned by the `runSimulation` method can be accessed with the method `getOutputValues`. The user can provide either the path(s) of the output (which can be a molecule, a parameter, or an observer),
  or the object(s) of the type `Molecule`, `Parameter`, or `Quantity` (for observers) with the argument `quantitiesOrPaths`. If no output is specified, all outputs available in the simulation results are returned.
  The paths of all available outputs can be accessed via

  ```{r getAllOutputSelections}
  populationResults$allQuantityPaths
  ```

  `getOutputValues` returns a list with two entries: `data` and `metadata`:

    * `data` is a dataframe with two predefined columns (IndividualId and Time) as well as one columne for each requested output

  * `IndividualId`
  * `Time` a vector with simulated time values (in minutes, equal for all outputs)
  *  a vector with simulated entries for each output requested.

  The values of `IndividualId`, `Time`, and the simulated outputs, are appended for each simulated invidual.  Note that this results in non-monotonously increasing column `Time`.

  ```{r getOutputValues}
  # Get simulated results by path
  resultsPath <- populationResults$allQuantityPaths[[1]]
  print(resultsPath)

  resultsData <- getOutputValues(populationResults, quantitiesOrPaths = resultsPath)

  resultsTime <- resultsData$data$Time
  resultsValues <- resultsData$data$`Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)`

  plot(resultsTime, resultsValues, type = "l")
  ```

  For more information about running simulations, please refer to [Running simulations and retrieving the results](run-simulation.html).
