---
title: "Sensitivity analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sensitivity analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.showtext = TRUE
)
```

The models built with OSPS depend on a lot of input parameters which are based on literature values, measurements, databases, and assumptions. For a given set of input parameters, a number of output curves is computed in a simulation. To assess which input parameters have most impact on the output curves, a sensitivity analysis of the simulation can be performed. For more information about theoretical background, see [OSPS documentation](https://docs.open-systems-pharmacology.org/shared-tools-and-example-workflows/sensitivity-analysis).

In brief, the values of the chosen parameters are changed by a certain percentage and the impact of these changes on PK parameters of model outputs is assessed.

## Performing a sensitivity analysis

The first step of performing a sensitivity analysis is creating an object of the class `SensitivityAnalysis`. At this step, the user can define which parameters should be considered for the sensitivity analysis, in which range the values are varied, and how may steps are performed in one direction (plus and minus). If no parameters are specified, all constant and suitable for sensitivity calculations parameters of the simulation will be varied. The list of such parameters paths for a simulation can be accessed with the method `potentialVariableParameterPathsFor(simulation)`.

### Understanding `numberOfSteps` and `variationRange`

The `numberOfSteps` parameter specifies the number of steps in one direction (either positive or negative) from the reference parameter value. Since parameters are varied in both directions, the total number of variations per parameter is `2 * numberOfSteps`.

The `variationRange` parameter defines the total relative range of variation (e.g., `0.1` means ±10%). The range is divided into equal intervals based on the number of steps.

**Example:** With `numberOfSteps = 2` and `variationRange = 0.1` (default values):
- Each parameter is tested at **four points**: 90%, 95%, 105%, and 110% of its reference value
- Total number of simulations: `2 * numberOfSteps * number_of_parameters`

**Example:** With `numberOfSteps = 1` and `variationRange = 0.1`:
- Each parameter is tested at **two points**: 90% and 110% of its reference value

```{r createSensitivityAnalysis}
library(ospsuite)

# Load simulation
simFilePath <- system.file("extdata", "simple.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)

# Get the paths of parameters that will be considered by default.
potentialSAParameters <- potentialVariableParameterPathsFor(sim)
print(potentialSAParameters)

# Create a default `SensitivityAnalysis` for the simulation
sa <- SensitivityAnalysis$new(simulation = sim)
print(sa)

# Create a `SensitivityAnalysis` with specified parameters
sa <- SensitivityAnalysis$new(simulation = sim, parameterPaths = c("Organism|Q", "Organism|Volume"))
print(sa)

# Show which parameters will be varied
sa$parameterPaths
```

New parameters can be added to an existing `SensitivityAnalysis` by calling the method `addParameterPaths()`.

**NOTE**: 
- If no parameters were specified during the creation of a `SensitivityAnalysis`, all constant and suitable for sensitivity calculation parameters are considered.
- In such cases, calling `addParameterPaths()` will only vary the newly added parameters.

```{r addPathsToSensitivityAnalysis}
library(ospsuite)

# Load simulation
simFilePath <- system.file("extdata", "simple.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)

# Create a `SensitivityAnalysis` with specified parameters
sa <- SensitivityAnalysis$new(simulation = sim, parameterPaths = c(
  "Organism|Q",
  "Organism|Volume"
))
# Add new parameter
sa$addParameterPaths("R1|k1")

# Show which parameters will be varied
sa$parameterPaths
```

To run the specified `SensitivityAnalysis`, call the method `runSensitivityAnalysis()`. The method returns an object of the class `SensitivityAnalysisResults`. The impact of the defined parameters is calculated for PK-Parameters (see [PK Analysis](pk-analysis.html) for more information) of all model outputs.

```{r runSA}
# Load simulation
simFilePath <- system.file("extdata", "simple.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)

# Create a `SensitivityAnalysis` with all constant and suitable for sensitivity calculations parameters and suitable for sensitivity calculations parameters
sa <- SensitivityAnalysis$new(simulation = sim)

# Run the sensitivity analysis
saResult <- runSensitivityAnalysis(sa)
print(saResult)
```

The method `allPKParameterSensitivitiesFor()` returns a list of `PKParameterSensitivity` objects. `PKParameterSensitivity` describes the sensitivity (field `$value`) of a PK-Parameter (`$pkParameterName`) for the output `$outputPath` calculated for the varied parameter `$parameterPath`. The argument `totalSensitivityThreshold` of the method `allPKParameterSensitivitiesFor()` is used to filter out the most impactful parameters. A threshold of `0.9` means that only parameters participating to a total of 90 percent of the sensitivity would be returned. A value of `1` would return the sensitivity for all parameters. If no value is provided, a default value is used.

### Customizing sensitivity analysis settings

The default values for `numberOfSteps`, `variationRange`, and `totalSensitivityThreshold` can be retrieved using `getOSPSuiteSetting("sensitivityAnalysisConfig")`. These values can be changed for a session by directly setting the values in `ospsuiteEnv`:

```r
# View current defaults
getOSPSuiteSetting("sensitivityAnalysisConfig")

# Change the defaults for the current R session
ospsuiteEnv$sensitivityAnalysisConfig$numberOfSteps <- 3
ospsuiteEnv$sensitivityAnalysisConfig$variationRange <- 0.2
ospsuiteEnv$sensitivityAnalysisConfig$totalSensitivityThreshold <- 0.95
```

Note that these changes only affect new `SensitivityAnalysis` objects created after the change. Existing objects retain their original values.

```{r allPKParameterSensitivitiesFor}
# Load simulation
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)

# Default value of the threshold
getOSPSuiteSetting("sensitivityAnalysisConfig")$totalSensitivityThreshold

# Create a `SensitivityAnalysis` with all constant parameters and run it
sa <- SensitivityAnalysis$new(
  simulation = sim,
  parameterPaths = c(
    "Aciclovir|Lipophilicity",
    "Aciclovir|Fraction unbound (plasma)",
    "Organism|Age"
  )
)
saResult <- runSensitivityAnalysis(sa)
print(saResult)

# Get sensitivities for the parameter "AUC_inf" of the simulated output with a threshold of 0.8
outputPath <- sim$outputSelections$allOutputs[[1]]$path
sensitivities <- saResult$allPKParameterSensitivitiesFor(pkParameterName = "AUC_inf", outputPath = outputPath, totalSensitivityThreshold = 0.8)
print(sensitivities)
```

The value `-1` for the sensitivity of "AUC_inf" of the output `Organism|PeripheralVenousBlood|Aciclovir|Plasma (Peripheral Venous Blood)` for the parameter `Organism-Kidney-Volume` means that increasing `Organism-Kidney-Volume` by 10% will decrease "AUC_inf" by 10%.
Note that the list of sensitivities is ordered from largest to smallest with respect to magnitude.

## Import and export of sensitivity analysis results

Sensitivity analysis calculated in R can be exported to a *.csv file, which can be loaded in another instance.

```{r SAExport}
# Load and run the simulation
simFilePath <- system.file("extdata", "simple.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)
simulationResults <- runSimulations(simulations = sim)[[1]]

# Create a `SensitivityAnalysis` with all constant parameters and run it
sa <- SensitivityAnalysis$new(simulation = sim)
saResult <- runSensitivityAnalysis(sa)

# Export to csv
saResultPath <- system.file("extdata", "SAResult.csv", package = "ospsuite")
exportSensitivityAnalysisResultsToCSV(results = saResult, filePath = saResultPath)

# Load from csv
saResultLoaded <- importSensitivityAnalysisResultsFromCSV(filePaths = saResultPath, simulation = sim)
```
