---
title: "Changing parameter and molecule start values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Changing parameter and molecule start values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", error = TRUE
)
```

## Changing initial values

It is possible to change the initial values of all parameters and molecules in a simulation. It is important to differentiate between the *constant* values and the values that are defined by a *formula*.

### Formula types

Every (initial) value is described either by a constant or by a formula. If the value is defined by a simple constant, the field `isConstant` of the respective parameter or molecule has the value `TRUE`.

```{r getConstantValue}
library(ospsuite)

# Load simulation
simFilePath <- system.file("extdata", "Aciclovir.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)

# Get the parameter "Age" of the Organism
ageParam <- getParameter("Organism|Age", sim)
print(ageParam)
ageParam$isConstant
```

If the value is not constant, it is described by one of the formula types:

  **Distributed**: Distributed parameters describe a variation around a constant value or between two numerical limits. In the simulation, the behavior equals that of constant values. See [OSPS documentation](https://docs.open-systems-pharmacology.org/working-with-mobi/mobi-documentation/model-building-components#working-with-constant-and-distributed-parameters) for more information.

```{r getDistributedValue}
# Get the parameter "Volume" of the Liver
liverVolume <- getParameter("Organism|Liver|Volume", sim)
print(liverVolume)
liverVolume$isDistributed
```

  **Explicit formula**: The value of the parameter is given by an explicit formula. The value of an explicit formula can change during the simulation. The string of the formula can be accessed via `parameter$formulaString`.

```{r getExplicitFormulaValue}
# Get the parameter "Volume" of the Liver interstital
liverIntVolume <- getParameter("Organism|Liver|Interstitial|Volume", sim)
print(liverIntVolume)
liverIntVolume$formulaString
```

  **Table formula**: The value of the parameter is given by a table with x-y value pairs. X values usually refer to the simulation time. See [Table parameters](table-parameters.html) for additional information on how to retrieve or change the table values.

```{r getTableFormulaValue}
# Get the parameter defined by a table.
tableParam <- getParameter("Organism|TableParameter", sim)
print(tableParam)
```

Additionally, some parameters are modeled as *state variables*. In this case, the parameter has an *initial value* and a *right hand side (RHS)* formula, both of which can be any formula type.

```{r getStateVariableValue}
# Get the parameter defined by a state variable.
stateVariableParam <- getParameter("Organism|StateVariable_Parameter", sim)
print(stateVariableParam)

# `value` refers to the initial value of the parameter
stateVariableParam$value
# `rhsFormula` is the right hand side of the parameter
stateVariableParam$rhsFormula
```

### Changing parameters and molecules initial values

The user can change initial values of parameters and molecules with the methods `setParameterValues` and `setMoleculeInitialValues`, respectively.

```{r changeDose}
# Get the parameter Dose
doseParamPath <- "Applications|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose"
doseParam <- getParameter(doseParamPath, sim)
print(doseParam)

# Change the dose to 350mg. The value has to be converted to base unit, first
newValue <- toBaseUnit(quantity = doseParam, values = 350, unit = "mg")
setParameterValues(parameters = doseParam, values = newValue)
print(doseParam)
```

Another way to change parameter values is to scale them. The scaling is always performed relative to the current value:

```{r scaleParameter}
doseParamPath <- "Applications|IV 250mg 10min|Application_1|ProtocolSchemaItem|Dose"
doseParam <- getParameter(doseParamPath, sim)
print(doseParam)

# Double the dose
scaleParameterValues(doseParam, factor = 2)
print(doseParam)

# Half the dose
scaleParameterValues(doseParam, factor = 0.5)
print(doseParam)
```

Only constant values can be set. If the parameters value is defined by a formula (explicit or table), the newly assigned value will override the formula. This will be reflected by the field `isFixedValue`. Changing values of formula parameters should be done with caution, as the potential dependency on another simulation parameters will be destroyed.

```{r setExplicitFormulaValue}
# Get the parameter "Volume" of the Liver interstital
liverIntVolume <- getParameter("Organism|Liver|Interstitial|Volume", sim)
print(liverIntVolume)

# isFixedValue is FALSE as the parameter is defined by its formula
liverIntVolume$isFixedValue

setParameterValues(liverIntVolume, 1)
print(liverIntVolume)
# isFixedValue is TRUE as the value of parameter is overridden by the constant value
liverIntVolume$isFixedValue
```

The parameter value can be reset to its formula after assigning a constant value:

```{r resetExplicitFormulaValue}
print(liverIntVolume)

# isFixedValue is TRUE as the value of parameter is overridden by the constant value
liverIntVolume$isFixedValue

liverIntVolume$reset()
print(liverIntVolume)
liverIntVolume$isFixedValue
```

Changing the value of a state variable parameter will only change its initial value, but not the right hand side (i.e., the value in the simulation will still change according to the RHS formula). To switch to RHS formula off and, by that, make the parameter be only dependent on its initial value, the field `isStateVariable` can be set to `FALSE`. This function is one-way only! It is not possible to re-activate the RHS formula after switching it off.

```{r setStateVariableValue, error = TRUE}
# Get the parameter defined by a state variable.
stateVariableParam <- getParameter("Organism|StateVariable_Parameter", sim)
print(stateVariableParam)

# Setting its value only changes the initial value
setParameterValues(stateVariableParam, 10)
print(stateVariableParam)

# Switching the RHS formula off
stateVariableParam$isStateVariable <- FALSE
print(stateVariableParam)

# Switching it on is not supported
stateVariableParam$isStateVariable <- TRUE
```

An example how to set the initial values of molecules in all containers to a certain value:

```{r changeInitialValue}
# Get objects representing the molecule Aciclovir in all containers
allAciclovirMolecules <- getAllMoleculesMatching("Organism|**|Aciclovir", sim)

# Set initial values to 10 µmol in all containers
setMoleculeInitialValues(allAciclovirMolecules, rep(10, length(allAciclovirMolecules)))
```
