---
title: "Working with snapshots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with snapshots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A *snapshot* is a JSON representation of a PK-Sim or MoBi project. Snapshots are
human-readable, diff-friendly, and version-control-tractable, which makes them a
convenient exchange format for projects. Read more about OSP snapshots in the [documentation](https://docs.open-systems-pharmacology.org/working-with-pk-sim/pk-sim-documentation/importing-exporting-project-data-models#exporting-project-to-snapshot-loading-project-from-snapshot). `{ospsuite}` exposes helpers to load
simulations from a snapshot, run simulations stored in a snapshot, and convert
between the snapshot and project formats.

```{r library, message=FALSE}
library(ospsuite)

# This vignette uses the example snapshot shipped with the package
snapshotPath <- system.file(
  "extdata",
  "test_snapshot.json",
  package = "ospsuite"
)
```

## Loading simulations from a snapshot

`loadSimulationsFromSnapshot()` creates the simulations stored in a snapshot file
and returns them as a named list of `Simulation` objects.

By default, every simulation in the snapshot is loaded:

```{r loadAll}
simulations <- loadSimulationsFromSnapshot(snapshotPath)

# Names of the loaded simulations
allNames <- names(simulations)
allNames
```

When you only need specific simulations, pass their names via `simulationNames`.
The match is case-sensitive:

```{r loadByName}
oneSimulation <- loadSimulationsFromSnapshot(
  snapshotPath,
  simulationNames = allNames[1]
)

oneSimulation[[1]]$name
```

By default, requesting a name that is not in the snapshot raises an error. Set
`ignoreIfNotFound = TRUE` to instead get a `NULL` entry for each missing name,
keeping the order of the requested names:

```{r loadNoMatch}
loadSimulationsFromSnapshot(
  snapshotPath,
  simulationNames = c(allNames[1], "Does not exist"),
  ignoreIfNotFound = TRUE
)
```

Once loaded, a simulation can be used like any other:

```{r useSimulation}
simulation <- simulations[[1]]

results <- runSimulations(simulation)[[1]]
results
```

## Running simulations from a snapshot

`runSimulationsFromSnapshot()` runs every simulation contained in one or more
snapshot files (or directories of snapshot files) and writes the results to an
output directory. Unlike `loadSimulationsFromSnapshot()`, it does not return the
simulations to R; instead it exports the results to disk in the requested
formats.

```{r runFromSnapshot}
outputDir <- file.path(tempdir(), "snapshot-results")
dir.create(outputDir, showWarnings = FALSE)

runSimulationsFromSnapshot(
  snapshotPath,
  output = outputDir,
  exportCSV = TRUE,
  exportPKML = TRUE
)

list.files(outputDir)
```

The export formats are controlled independently:

* `exportCSV` (default `TRUE`) — results as CSV.
* `exportPKML` (default `FALSE`) — simulations as PKML.
* `exportJSON` (default `FALSE`) — results as JSON.
* `exportXML` (default `FALSE`) — simulations as XML.

Set `RunForAllOutputs = TRUE` to compute all model outputs instead of only the
output selections defined in the snapshot.

You can also point the function at a directory and it will process every snapshot
in it:

```{r runDirectory}
snapshotDir <- file.path(tempdir(), "snapshots")
dir.create(snapshotDir, showWarnings = FALSE)
file.copy(snapshotPath, snapshotDir, overwrite = TRUE)

runSimulationsFromSnapshot(snapshotDir, output = outputDir)
```

## Converting between snapshot and project formats

Snapshots and PK-Sim project files (`.pksim5`) can be converted into each other.
Use `loadProjectFromSnapshot()` to build a project from a snapshot, and
`exportProjectToSnapshot()` to write a project back out as a snapshot.

Convert a snapshot to a PK-Sim project:

```{r loadProjectFromSnapshot}
projectDir <- file.path(tempdir(), "project")
dir.create(projectDir, showWarnings = FALSE)

loadProjectFromSnapshot(snapshotPath, output = projectDir)

projectFile <- list.files(projectDir, pattern = "\\.pksim5$", full.names = TRUE)
projectFile
```

Set `runSimulations = TRUE` to also run the simulations during this conversion.

And convert a project back to a snapshot:

```{r exportProjectToSnapshot}
snapshotOut <- file.path(tempdir(), "snapshot")
dir.create(snapshotOut, showWarnings = FALSE)

exportProjectToSnapshot(projectFile, output = snapshotOut)

list.files(snapshotOut)
```

As with `runSimulationsFromSnapshot()`, both functions accept several files or
directories at once.

> **Note**
>
> `convertSnapshot()` is deprecated in favour of `loadProjectFromSnapshot()` and
> `exportProjectToSnapshot()`.
