---
title: "Working with snapshots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with snapshots}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

A *snapshot* is a JSON representation of a PK-Sim or MoBi project. Snapshots are
human-readable, diff-friendly, and version-control-tractable, which makes them a
convenient exchange format for projects. `{ospsuite}` exposes helpers to load
simulations from a snapshot, run simulations stored in a snapshot, and convert
between the snapshot and project formats.

All snapshot helpers require PK-Sim to be installed and initialized. The
functions take care of initialization for you, but make sure PK-Sim is available
(see `vignette("PKSim-installation")`).

```{r library}
library(ospsuite)
```

## Loading simulations from a snapshot

`loadSimulationsFromSnapshot()` reads the simulations stored in a snapshot file
and returns them as a list of `Simulation` objects. These behave exactly like
simulations loaded from a `.pkml` file with `loadSimulation()`, so you can
inspect their parameters, modify them, and run them.

By default, every simulation in the snapshot is loaded:

```{r loadAll}
simulations <- loadSimulationsFromSnapshot("path/to/snapshot.json")

# Names of the loaded simulations
vapply(simulations, function(sim) sim$name, character(1))
```

When you only need specific simulations, pass their names via `simulationNames`.
The match is case-sensitive:

```{r loadByName}
simulations <- loadSimulationsFromSnapshot(
  "path/to/snapshot.json",
  simulationNames = c("Simulation A", "Simulation B")
)
```

If no simulation in the snapshot matches the requested names, an empty list is
returned.

Once loaded, a simulation can be used like any other:

```{r useSimulation}
simulation <- loadSimulationsFromSnapshot("path/to/snapshot.json")[[1]]

results <- runSimulations(simulation)[[1]]
```

> **Note**
>
> `loadSimulationsFromSnapshot()` is **experimental** and depends on a
> computational-core follow-up
> ([PK-Sim#3560](https://github.com/Open-Systems-Pharmacology/PK-Sim/issues/3560)):
> snapshot-loaded simulations must carry a simulation configuration before they
> can be wrapped into `Simulation` objects. Until that ships, this function
> errors. To *run* simulations stored in a snapshot in the meantime, use
> `runSimulationsFromSnapshot()` (see below).

## Running simulations from a snapshot

`runSimulationsFromSnapshot()` runs every simulation contained in one or more
snapshot files (or directories of snapshot files) and writes the results to an
output directory. Unlike `loadSimulationsFromSnapshot()`, it does not return the
simulations to R; instead it exports the results to disk in the requested
formats.

```{r runFromSnapshot}
runSimulationsFromSnapshot(
  "path/to/snapshot.json",
  output = "path/to/output",
  exportCSV = TRUE,
  exportPKML = TRUE
)
```

The export formats are controlled independently:

* `exportCSV` (default `TRUE`) — results as CSV.
* `exportPKML` (default `FALSE`) — results as PKML.
* `exportJSON` (default `FALSE`) — results as JSON.
* `exportXML` (default `FALSE`) — results as XML.

Set `RunForAllOutputs = TRUE` to compute all model outputs instead of only the
output selections defined in the snapshot.

You can pass several snapshot files or directories at once:

```{r runMultiple}
runSimulationsFromSnapshot(
  "path/to/snapshot1.json",
  "path/to/folder_of_snapshots",
  output = "path/to/output"
)
```

## Converting between snapshot and project formats

`convertSnapshot()` converts files between the snapshot (`.json`) and project
(`.pksim5`) formats. The `format` argument names the *target* format:

```{r convertToProject}
# Snapshot -> project
convertSnapshot(
  "path/to/snapshot.json",
  format = "project",
  output = "path/to/output"
)

# Project -> snapshot
convertSnapshot(
  "path/to/project.pksim5",
  format = "snapshot",
  output = "path/to/output"
)
```

When converting from a snapshot to a project, set `runSimulations = TRUE` to also
run the simulations during conversion.

As with `runSimulationsFromSnapshot()`, you can pass several files or directories
at once.
