---
title: "Table parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Table parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# Load the shared runtime gate; see vignettes/_setup.R.
source("_setup.R")
knitr::opts_chunk$set(
  eval = .ospRuntimeAvailable,
  collapse = TRUE,
  comment = "#>",
  fig.showtext = TRUE
)
```

Parameters defined by a **table formula** have special syntax for retrieving and changing the values. The value of a table parameter is given by a list of `ValuePoint` objects, each `ValuePoint` being an x-y value pair. X values usually refer to the simulation time.

```{r getTableFormulaValue}
library(ospsuite)

# Load a simulation that contains a table parameter
simFilePath <- system.file("extdata", "simple.pkml", package = "ospsuite")
sim <- loadSimulation(simFilePath)

# Get the parameter defined by a table.
tableParam <- getParameter("Organism|TableParameter", sim)
print(tableParam)
```

Direct access to the value points is possible through the `TableFormula` of the table parameter. All x- or y-values stored in the table can be conveniently retrieved using the `lapply` method:

```{r getAllXYValues}
# Get the parameter defined by a table
tableParam <- getParameter("Organism|TableParameter", sim)
print(tableParam)

# Get all value points
tableParam$formula$allPoints

# Get all x-values
xValues <- lapply(tableParam$formula$allPoints, function(point) {
  point$x
})
print(xValues)

# Get all y-values
yValues <- lapply(tableParam$formula$allPoints, function(point) {
  point$y
})
print(yValues)
```

The method `valueAt()` of the `TableFormula` returns the value of `y` for the given `x`. If no entry exists for the `x`, the value `y` is linearly interpolated between the two closest `x` values.

```{r valueAt}
# Get the parameter defined by a table
tableParam <- getParameter("Organism|TableParameter", sim)

# Value at x = 30 is stored in the table
tableParam$formula$valueAt(30)

# Value at x = 20 is not in the table
tableParam$formula$valueAt(20)
```

## Changing table parameter values
Simply setting the value of a table-defined parameter using `setParameterValues` will override the formula and make the parameter constant.

```{r setTableToConstant}
# Get the parameter defined by a table.
tableParam <- getParameter("Organism|TableParameter", sim)
tableParam

# Set value to a constant. tableParam$isFixedValue is now TRUE
setParameterValues(tableParam, 10)
tableParam
```

To change the values of the table, a set of methods of the `TableFormula` is available. The method `addPoints()` adds a set of x-y values to the existing table. If trying to add a point with an x-value that is already present in the table and a different y-value, an error is thrown:

```{r addPoints, error = TRUE, purl = FALSE}
tableParam <- getParameter("Organism|TableParameter", sim)
tableParam

# Add new points
tableParam$formula$addPoints(c(1, 2, 3), c(5, 6, 7))
tableParam

# Try to add points with existing x-values
tableParam$formula$addPoints(0, 9)
```

To remove a point from the table, use the method `removePoint()`. It remove a point if the x value is present in the table and has the provided y.

```{r removePoint}
tableParam

# Remove the point (0, 1)
tableParam$formula$removePoint(0, 1)
tableParam

# Try to remove the point (1, 1). Note that the y value for x = 1 is 5 in the table,
# and no point is removed.
tableParam$formula$removePoint(1, 1)
tableParam

# Try to remove the point (0, 1) again. It is no longer in the table, and no
# point is removed.
tableParam$formula$removePoint(0, 1)
tableParam
```

The `clearPoints()` method removes all points from the table, while `setPoints()` method is a combination of clearing the table and adding new points:

```{r setPoints}
tableParam

tableParam$formula$setPoints(c(1, 2, 3, 4), c(5, 6, 7, 8))
tableParam

tableParam$formula$clearPoints()
tableParam
```
